---
title: "Anomaly detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anomaly detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Two functions ask different questions, and the difference is the whole point.

```{r}
library(illume)
```

## One column, or the combination

`ilm_outliers()` asks whether a **value** is extreme for its own column.

```{r}
ilm_outliers(d, "height")
ilm_outliers_all(d)            # every column at once
```

`ilm_anomaly()` asks whether a **row** is a plausible combination.

```{r}
ilm_anomaly(d)
```

Someone 150 cm tall is unremarkable. Someone weighing 110 kg is unremarkable.
Someone who is both is not, and nothing in either column's distribution says so.
On simulated data with ten rows pushed off the correlation structure but inside
every column's range, the column-at-a-time scan caught **none** of them.

The method: a set of correlated columns puts most of its variation in a few
directions. Fit those directions, project each row onto them, and score how much
of the row is left over. A row that respects the correlations needs only a few
numbers to describe it; one that does not leaves a large remainder.

```{r}
r <- ilm_anomaly(d)
head(r)
#>  row score        p  p_adj  flag driver
#>    1 3.746 8.55e-05 0.0256  TRUE      a
```

`driver` is the column contributing most to that row's score. An anomaly nobody
can explain is not actionable, so it is reported alongside the score.

## Three things that had to be measured

This looks like a short function and is not. Two of the three decisions below
overturned the first version of it.

### The rank is not the imputation rank

`ilm_impute()` chooses the number of directions by cross-validating held-out
cells, which is exactly right for filling a value in and wrong here. On a rank-2
structure in eight columns it chose **6 or 7** -- on clean data as well as
contaminated -- and those extra components span the very directions the
anomalies depart along. Detection fell from 0.975 to 0.560.

`ilm_anomaly()` uses parallel analysis instead: permute each column
independently, which destroys everything the columns share while leaving each
one's own distribution alone, and keep a component only when it beats what the
permuted data produces. That chose 2 on every dataset tried, contaminated or
not.

The two functions answer different questions -- what best predicts a missing
cell, against how many directions are real shared structure -- so they use
different criteria, deliberately.

### The fit is trimmed

The anomalies sit in the same data the directions are estimated from, so they
pull those directions towards themselves and are then reconstructed well. The
method hides what it is looking for.

`trim = 0.25` fits on the best-fitting three quarters of rows and refits, so a
row is scored against a structure it did not help define. It earns its place
only where the anomalies share a direction, which is where they can form a
component between them:

```
  anomalies along random directions      along ONE shared direction
   2%   0.925  vs  0.917                  2%   0.950  vs  0.942
   5%   0.967  vs  0.977                  5%   0.797  vs  0.887
  10%   0.970  vs  0.978                 10%   0.587  vs  0.670
```

Holding rows out in folds instead -- which was the first design -- does nothing,
and is not offered. At four contamination levels it matched in-sample scoring to
three decimals, because four fifths of the anomalies remain in every training
fold.

### The reference is simulated, and calibrated

The score is not chi-squared: the noise scale is estimated, the rank came from
the same data, and the residual is taken against an estimated subspace. Datasets
with the same structure and no anomalies are simulated and scored through the
identical pipeline.

Getting that simulation right mattered twice. The noise has to be **per column**
-- standardising gives every column unit variance, so one that loads weakly on
the shared directions keeps proportionally more of its variance off them -- and
its level has to be matched to the observed median rather than read off shrunken
in-sample residuals. A first version got both wrong and flagged **38.9%** of the
rows of data containing no anomalies at all.

As it stands: across 100 clean datasets of 400 rows, 11 rows in 40,000 were
flagged, with 95 of the 100 producing none; with anomalies present, 98% of the
planted rows were found at 2% contamination and 99.7% at 5%.

`B` defaults to 39 because a Benjamini-Hochberg adjusted p-value cannot fall
much below `1 / B`. At `B = 19` a lone anomaly among 300 rows could not be
flagged however extreme it was, and the function says so rather than returning
an empty list.

## What it cannot do

When a large share of rows depart along the **same** direction, they are not
anomalies -- they are a subpopulation, and a rank-`k` fit of the whole data
legitimately includes their direction. Detection degrades accordingly: 0.89 at
5% contamination, 0.67 at 10%, 0.41 at 20%.

That is the method reaching its limit rather than failing quietly, and the tool
for a second group is `ilm_cluster()`, which is looking for exactly that.

```{r}
ilm_profile(d)                 # is it a group rather than a scattering?
```

## A flagged row is not an error

It is a combination the other rows do not make. It may be a data-entry mistake,
or a real and interesting case, or the most important observation in the
dataset. Deleting it because a method said so is how real effects get removed,
and the print says so every time.

## See also

`?ilm_anomaly`, `?ilm_outliers`, `vignette("profiling")` for the dimension
reduction underneath, and `vignette("workflow")` for where this sits.
