---
title: "ANOVA: factorial and repeated measures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ANOVA: factorial and repeated measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
library(illume)
```

If you have run an ANOVA in SPSS, or used `afex` in R, this will look familiar.
You describe the design by saying what each column *is* -- who the participant
is, what was measured, what varied between people and what varied within them
-- and you get back the table you were expecting, with the sphericity
correction and the effect size already in it.

```{r}
library(illume)
```

## The design we will use

Thirty-six people, randomised to one of three doses, tested three times. Dose
varies **between** participants (each person has one dose). Session varies
**within** participants (each person is measured at all three). `baseline` is a
score taken once, before anything happened.

```{r}
head(d, 4)
#>   id session    dose   recall baseline
#> 1  1     pre placebo 51.48496      113
#> 2  2     pre placebo 49.92847      111
#> 3  3     pre placebo 49.19631      126
#> 4  4     pre placebo 50.27987      102
```

One row per person per session. This is "long" format, and it is the format you
want -- nothing needs reshaping into columns.

## The analysis

```{r}
a <- ilm_aov_ez(id = "id", dv = "recall", data = d,
                between = "dose", within = "session")
a
```

```
Analysis of Variance
  recall ~ dose (between) + session (within)
  36 participants

       Effect    df   MSE      F   ges       p
         dose 2, 33 6.227 10.591 0.298 0.00028 ***
      session 2, 66 1.604 62.484 0.392 < 1e-04 ***
 dose:session 4, 66 1.604 15.641 0.244 < 1e-04 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Sphericity (Mauchly):
       Effect     W     p    GG HF
      session 0.998 0.972 0.998  1
 dose:session 0.998 0.972 0.998  1
```

That is the whole thing. No `Error()` term, no `aov()`, no reshaping.

### Reading it across

**`df`** is two numbers: how many ways the effect can vary, then how much
information is left to estimate the noise. `2, 33` for dose means three dose
groups (two differences between them) tested against 33 participants' worth of
residual.

Notice that `session` is tested on **66** and `dose` on **33**. They are
different because they are different comparisons. A within-participant effect
gets to use each person as their own control, so it has more information behind
it -- which is exactly why repeated-measures designs are efficient.

**`MSE`** is the noise those F tests are dividing by. It is smaller for the
within-participant effects (1.604 against 6.227) for the same reason.

**`F`** is signal over noise. **`p`** is how often you would see an F that large
if nothing were going on.

**`ges`** is the effect size, and it is worth a section of its own.

## Why `ges` and not eta squared

You have probably met three effect sizes for ANOVA and been told to report one:

* **eta squared** is this effect's variance over *all* the variance. It shrinks
  whenever you add anything else to the design, so the same effect looks smaller
  in a bigger study.
* **partial eta squared** is this effect's variance over itself plus its own
  error. It does not shrink -- but it *inflates* in repeated-measures designs,
  because the between-participant variation has been removed from its
  denominator. A within-participant effect and a between-participant effect
  reported this way are not on the same scale, and neither is comparable to a
  between-participants study of the same thing.
* **generalized eta squared (`ges`)** puts the variability back into the
  denominator that a different design would have had. The result is comparable
  across designs, which is what you need if your number is going into a
  meta-analysis or a power calculation for someone else's study.

`ges` is the default here for the same reason it is the default in `afex`.
Olejnik and Algina (2003) is the source; Bakeman (2005) is the readable version.

One thing `ges` needs from you. Its denominator should include variables that
were **measured** rather than **manipulated** -- age, baseline score, anything
you recorded but did not assign. Name those in `observed`:

```{r}
ilm_aov_ez("id", "recall", d, between = "dose", within = "session",
           observed = "dose")     # if dose had been self-selected, not assigned
```

Covariates are measured by definition, so they are added for you.

## Sphericity, and what to do about it

The F test for a within-participant effect assumes **sphericity**: that the
differences between every pair of sessions are equally variable. With only two
levels there is one difference and nothing to violate. With three or more, it
can fail -- and when it does, the test says "significant" more often than it
should.

`ilm_aov_ez()` runs Mauchly's test and reports it, along with the two standard
corrections:

```
Sphericity (Mauchly):
       Effect     W     p    GG HF
      session 0.998 0.972 0.998  1
 dose:session 0.998 0.972 0.998  1
```

`W` near 1 and a large `p` mean sphericity is fine here. `GG` and `HF` are
**epsilon** values: how far from spherical things are, where 1 is perfectly
spherical. They shrink the degrees of freedom, which makes the test more
conservative.

By default the correction is applied only when Mauchly's test rejects. You can
force it:

```{r}
ilm_aov_ez("id", "recall", d, between = "dose", within = "session",
           correction = "GG")     # or "HF", or "none"
```

### The thing nobody tells you about corrections

A correction is a patch on a test whose assumption has failed. There is another
option: use a model that never assumed sphericity in the first place.

```{r}
ilm_model(recall ~ dose * session + (1 | id), data = d,
          re_struct = "us", family = "gaussian")
```

`re_struct = "us"` leaves the covariance between sessions completely free, so
there is nothing to correct. If sphericity fails badly, `ilm_aov_ez()` will say
so and point you here. It is not always the right move -- it costs parameters,
and with few participants that matters -- but it is the honest alternative to
adjusting a broken test.

## Following up a significant effect

An F test with three or more levels tells you the means are not all the same. It
does not tell you which ones differ, and that is usually the finding.

`ilm_aov_ez()` does this for you, for significant effects where there is
something to compare. (A two-level factor is skipped: its F test *is* the
comparison.)

For a **main effect**, it is the pairwise comparisons. For an **interaction**,
it is something more useful.

### Interactions get simple effects, not every cell

A 3 × 3 interaction has nine cells and thirty-six possible pairs. Most of those
pairs change both factors at once, and answer nothing you asked.

What takes an interaction apart is the **simple effect**: hold one factor still,
compare the other. In both directions, because they are different questions.

```{r}
a$posthoc[["dose:session"]]
```

```
  ~ dose within each session ~
    [session = pre]
       contrast estimate    se  lower upper  p_adj
  low - placebo  -0.6298 0.724 -2.330 1.071 0.6625
 high - placebo  -0.3525 0.724 -2.053 1.348 0.8810
     high - low   0.2773 0.724 -1.423 1.978 0.9220

    [session = mid]
       contrast estimate    se   lower upper   p_adj
  low - placebo   0.5744 0.724 -1.1268 2.276 0.71235
 high - placebo   2.6469 0.724  0.9457 4.348 0.00045
     high - low   2.0725 0.724  0.3714 3.774 0.01115

    [session = post]
       contrast estimate    se   lower upper   p_adj
  low - placebo    1.598 0.724 -0.1028 3.299 0.07065
 high - placebo    5.382 0.724  3.6810 7.083 0.00000
     high - low    3.784 0.724  2.0828 5.485 0.00000
```

Read down the sessions and the finding writes itself: at **pre** the groups are
indistinguishable, which is what randomisation should give you; by **mid** the
high dose has separated from placebo; by **post** the separation is large and
the doses are ordered. That is a dose-response emerging over time, and it is
what the interaction was pointing at.

The other direction -- how each dose group changes across sessions -- is in the
same object, below.

### "Significant in one group but not the other" is not an interaction

This is the most common mistake in this part of a paper. If the drug group
improves significantly and the placebo group does not, that is **two separate
tests**, and it does not establish that the groups differ from each other. The
difference between a significant result and a non-significant one is not itself
significant.

The test you need is the one comparing the groups -- the `~ dose within each
session ~` half above, or the interaction F itself. Both are printed, and they
are printed together so that the comparison is in front of you.

## When your predictor is continuous

Sometimes time is not three sessions but a number -- week 0, 2, 6, 12 -- and you
want to know whether the *slope* differs by group. ANOVA cannot do that: it
needs levels. This is what `ilm_trends()` is for.

```{r}
f <- ilm_model(recall ~ dose * week + (1 | id), data = d,
               family = "gaussian", reml = TRUE)
ilm_trends(f, "dose", var = "week")
```

```
Estimated marginal slopes of `week` within dose
  averaged with equal weights; t tests on satterthwaite df
    dose estimate     se      df statistic p.value   lower  upper
 placebo   0.3398 0.2534 69.0000    1.3409  0.1844 -0.1658 0.8454
     low   1.4538 0.2534 69.0001    5.7366  0.0000  0.9482 1.9594
    high   3.2071 0.2534 69.0001   12.6547  0.0000  2.7015 3.7127

  p-values above test each slope against ZERO.
```

That answers "is anything happening in this arm?". For "do the arms differ from
each other?", which is the different question, pass it on:

```{r}
ilm_contrast(ilm_trends(f, "dose", var = "week"))
```

```
       contrast estimate     se  lower upper  p_adj
  low - placebo    1.114 0.3584 0.2715 1.956 0.0045
 high - placebo    2.867 0.3584 2.0248 3.710 0.0000
     high - low    1.753 0.3584 0.9108 2.596 0.0000
```

Both are reported because both get asked, and they can disagree in either
direction.

## Covariates

A covariate is a numeric variable you want to adjust for. Add it by name:

```{r}
ilm_aov_ez("id", "recall", d, between = "dose", within = "session",
           covariate = "baseline")
```

```
           effect num_df den_df    F_value         ges      p_value
             dose      2     32 11.1525017 0.307451575 2.113305e-04
         baseline      1     32  1.1591189 0.023070248 2.896984e-01
          session      2     64  1.2036463 0.012441974 3.067993e-01
     dose:session      4     64 15.0581990 0.239678082 1.000167e-08
 baseline:session      2     64  0.4850469 0.005077051 6.179131e-01
```

### Your covariate has to be one number per person

`baseline` was measured once, so each participant has one value. That is what a
classical ANCOVA can use.

If your covariate was measured *at every session* -- daily stress, sleep the
night before, session-specific dose -- `ilm_aov_ez()` will stop and tell you so.
This is not fussiness. A time-varying covariate carries two effects at once:

* how people who are **generally** more stressed differ from people who are not;
* what happens when a person is **more stressed than usual for them**.

Those can be different sizes, and they can point in opposite directions. A
single coefficient gives you a blend of the two that answers neither question,
and the classical ANOVA has nowhere to put the distinction. The mixed model
does, by splitting the covariate into a person-average and a deviation from it.

## When people are missing a session

Classical repeated-measures ANOVA cannot use a participant who missed one
session. Not "handles it badly" -- cannot use them at all, because every
within-participant comparison needs all their cells. They get dropped.

`ilm_aov_ez()` tells you when that happens:

```
ilm_aov_ez: 35 participants, 3 within-participant cell(s)
  1 participant(s) dropped for incomplete cells -- classical repeated measures
  cannot use a participant missing an occasion. engine = "mixed" keeps them.
```

If you lost one person out of thirty-six, that is a footnote. If you lost
fifteen, the analysis is being done on a self-selected subset -- the people who
came back -- and that is a different study from the one you ran. A mixed model
uses everyone's available data:

```{r}
ilm_aov_ez("id", "recall", d, between = "dose", within = "session",
           engine = "mixed")
```

That translates your specification into the right model formula, which you can
see and check:

```{r}
a$formula
#> recall ~ dose * session + (1 | id)
```

## Coming from afex or SPSS

| You know | Here |
|---|---|
| `aov_ez(id, dv, data, between, within)` | `ilm_aov_ez(id, dv, data, between, within)` |
| `anova_table = list(es = "ges")` | reported by default |
| `correction = "GG"` | `correction = "GG"` |
| `observed = "age"` | `observed = "age"`; covariates are automatic |
| `emmeans(fit, ~ a)` then `pairs()` | done for you; or `ilm_emmeans()` + `ilm_contrast()` |
| `emtrends()` | `ilm_trends()` |
| SPSS "Tests of Within-Subjects Effects" | the same table, with `ges` instead of partial eta squared |

The numbers agree with `afex` to machine precision -- F to twelve decimal
places, `ges` to fourteen, and the Greenhouse-Geisser fractional degrees of
freedom to sixteen. If you have an existing analysis, you can check it.

## What this does not do

It does not decide anything for you. It will tell you that sphericity failed,
that eleven participants were dropped, or that your covariate varies within
person -- and in each case name what to do instead. Whether the classical
analysis is still the right one for your question is yours to judge, and the
notes at the bottom of the output are there so that it is judged rather than
missed.

## See also

`?ilm_aov_ez`, `?ilm_trends`, `?ilm_emmeans`, `?ilm_contrast`,
`vignette("regression-models")` for the general modelling engine underneath,
and `vignette("effect-size-and-power")` for planning the next study.

Olejnik, S. and Algina, J. (2003). Generalized eta and omega squared
statistics. *Psychological Methods* 8, 434-447.

Bakeman, R. (2005). Recommended effect size statistics for repeated measures
designs. *Behavior Research Methods* 37, 379-384.
