---
title: "Causal models: DAGs, difference in differences, regression discontinuity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal models: DAGs, difference in differences, regression discontinuity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
set.seed(2024)
library(illume)
```

A regression coefficient is an association. Turning one into a causal claim
takes an argument that lives outside the data, and this vignette covers the
three illume supports: a causal graph you supply, a difference in differences,
and a regression discontinuity.

What they share is that each names an assumption, and each lets that assumption
be checked at least partly. The functions here put the check in the same object
as the estimate, so it is hard to report one without the other.

## The one rule

**The design fixes which regression is fitted, and nothing searches over it.**

That matters more than it might sound. If you try specifications until the
diagnostics look good and then report the winner's p-values, those p-values are
no longer what they claim: you have used the data twice, once to choose and once
to test. illume's coverage is nominal across 23 designs precisely because
nothing does that.

So `ilm_dag_model()` takes the mean structure from the graph and leaves it
alone. What it *will* adjust, when a diagnostic asks, is the **error**
structure -- a model for the dispersion, a random effect for grouping. Those
change what the standard errors are, not which effect is being estimated. Every
step is recorded and printed as it happens.

## Part 1: a causal graph

### Writing one down

`ilm_dag()` reads the `dagitty` text form, an edge list, or a `dagitty` object.

```{r}
g <- ilm_dag("dag {
  treatment [exposure]
  recovery  [outcome]
  severity -> treatment
  severity -> recovery
  treatment -> recovery
  age -> severity
  age -> recovery
}")
g
```

The graph algorithms are written in illume rather than taken from `dagitty`,
which imports V8 -- a JavaScript engine is a heavy thing to require of someone
who wants to fit a regression. `dagitty` is in `Suggests` and the test suite
pins these functions to it: across 400 random graphs, 1564 d-separation tests
with no disagreements, 391 of 391 identical adjustment sets, and 2532 implied
claims all confirmed.

### What has to be adjusted for

```{r}
ilm_adjust_sets(g)
```

`severity` closes the back door. `age` does not need to be in the set as well,
because conditioning on `severity` already blocks the path through it.

The interesting case is when nothing works:

```{r}
ilm_adjust_sets(g, observed = c("treatment", "recovery", "age"))
```

An empty result is a **finding**. It says that with these variables measured, no
regression estimates this effect. That is the most useful thing a graph can tell
you, and it can only be said before the modelling rather than after.

### What the graph claims about the data

Every missing arrow is a testable claim. `ilm_dag_implied()` lists them and
`ilm_dag_test()` checks them:

```{r}
ilm_dag_implied(g)
```

The conditioning set is minimal -- no member can be dropped. Conditioning on
both variables' parents always works and is the usual textbook basis, but it is
often much larger than it needs to be, and every unnecessary covariate costs
power and invites a modelling error of its own.

```{r}
n <- 800
age      <- rnorm(n)
severity <- 0.7 * age + rnorm(n)
treatment <- rbinom(n, 1, plogis(0.8 * severity))
recovery <- 1.2 * treatment - 0.9 * severity + 0.5 * age + rnorm(n)
d <- data.frame(age, severity, treatment, recovery)

ilm_dag_test(g, d)
```

A failed claim does not say which side is at fault. The graph may be missing an
arrow, or a variable may not measure what its name supposes. Both are worth
knowing before any estimate is taken seriously. Note the two thresholds: a claim
counts as contradicted only if it is both significant after adjustment *and*
larger than `min_effect`. With a few thousand rows a correlation of 0.03 is
significant and means nothing.

### The whole workflow

```{r}
fit <- ilm_dag_model(g, d)
```

```{r}
fit
```

Compare that with ignoring the graph entirely:

```{r}
coef(ilm_model(recovery ~ treatment, data = d, family = "gaussian",
               verbose = FALSE))["treatment"]
```

The truth is 1.2. Adjusting for what the graph says to adjust for recovers it;
not adjusting does not.

### Several adjustment sets are a sensitivity analysis

When more than one minimal set is admissible, all of them are fitted. They
target the same quantity, so if the graph is right they should agree.

```{r}
g2 <- ilm_dag("dag {
  x [exposure] ; y [outcome]
  u -> a -> x ; u -> b -> y ; x -> y
}")
ilm_adjust_sets(g2)
```

The back-door path can be blocked at `a`, at `u`, or at `b`. Agreement across
the three supports the graph; a set that disagrees markedly is worth more
attention than any single point estimate.

### Reading it back in words

```{r}
ilm_interpret(fit, ame = FALSE)
```

Note that this says "leads to" rather than "is associated with", and says why.
Without a design that identifies an effect, `ilm_interpret()` will not use
causal language at all -- and it tells you that the licence rests on the graph
being right, which is an assumption you supplied rather than something the data
established.

## Part 2: difference in differences

Some units get treated at a point in time, others do not, and the comparison is
between the two groups' *changes* rather than their levels.

```{r}
panel <- expand.grid(unit = 1:40, time = 1:8)
panel$treated <- as.integer(panel$unit <= 20)
panel$post    <- as.integer(panel$time >= 5)
panel$y <- 1 + 0.3 * panel$time + rnorm(40)[panel$unit] +
  0.8 * panel$treated * panel$post + rnorm(nrow(panel))

did <- ilm_did(panel, "y", "unit", "time", treated = "treated", post = "post")
```

```{r}
did
```

The estimate is an interaction in a mixed model, so the fit is an ordinary
`ilm_model` and every method and diagnostic in the package applies to it.

### Parallel trends

The identifying assumption is that without the treatment the two groups would
have moved together. That cannot be checked where it matters -- it concerns a
world that did not happen -- but it can be checked *before* the treatment, and
the event study shows the shape of any departure:

```{r}
ilm_plot_did(did)
```

Pre-treatment coefficients should sit near zero. Here is what a violation looks
like:

```{r}
bad <- panel
bad$y <- bad$y + 0.35 * bad$treated * bad$time   # already diverging
did_bad <- ilm_did(bad, "y", "unit", "time", treated = "treated",
                   post = "post", verbose = FALSE)
did_bad$parallel$status
ilm_plot_did(did_bad)
```

Getting that check calibrated took choosing the right reference distribution.
The comparison is between two groups of *units'* slopes, so units are the
independent replicates and the large-sample normal is too generous when there
are few: it rejected 6.0%, 6.8% and 5.0% of the time at 40, 20 and 80 units,
where `t` on `units - 2` gave 4.8%, 5.5% and 4.5%. A check that cries wolf is
worse than no check, because the remedy for a failed parallel-trends test is to
abandon the design.

### Staggered adoption is refused

When units are treated at different times, a pooled two-way fixed effects
estimate is **not** an average treatment effect if the effect varies:
already-treated units end up serving as controls for later-treated ones.

```{r, error = TRUE}
stag <- panel
start <- ifelse(stag$unit <= 10, 4, ifelse(stag$unit <= 20, 6, Inf))
stag$treatment <- as.integer(stag$time >= start)
ilm_did(stag, "y", "unit", "time", treatment = "treatment", verbose = FALSE)
```

`allow_staggered = TRUE` returns the estimate with the caveat attached. The
honest remedy is a cohort-based estimator -- the `did` and `fixest` packages
implement them -- and illume does not have one.

### Serial correlation

Repeated observations of the same unit are correlated, and difference in
differences with naive standard errors over-rejects badly as a result. A unit
random intercept is fitted by default; `ar = TRUE` adds AR(1) on top, which is
what matters when the correlation decays rather than being constant.

## Part 3: regression discontinuity

Treatment is assigned by whether a running variable crosses a cutoff, so units
just either side are comparable and the jump at the cutoff is the effect.

```{r}
n <- 2000
run <- runif(n, -1, 1)
score <- 0.5 * run + 0.8 * (run >= 0) + rnorm(n, 0, 0.5)
rd <- ilm_rdd(data.frame(run = run, score = score), "score", "run", cutoff = 0)
```

```{r}
rd
ilm_plot_rdd(rd)
```

### The design checks

A regression discontinuity is only as good as the claim that crossing the cutoff
is the *only* thing that changes there. Four things check that.

**Density.** If people can move themselves across the cutoff, the ones just
above are not comparable to the ones just below.

```{r}
rd$density$status
```

The test is whether the density is *discontinuous* at the cutoff, not whether it
is *symmetric* about it -- and the difference is not academic. Comparing raw
counts either side against a 50/50 split flagged 76% of a perfectly smooth
sloped normal and 100% of a smooth exponential, over 500 unmanipulated data
sets. A local linear density fitted on each side and compared at the cutoff held
between 4.6% and 6.0% across uniform, normal and exponential running variables,
and had more power as well: with 30% of the units just below the cutoff moved
above it, the count split caught 81% and the local linear fit 98%.

**Covariate balance.** Anything measured before treatment should not jump.

```{r}
covs <- data.frame(run = run, score = score,
                   prior = rnorm(n), leaky = rnorm(n) + 0.9 * (run >= 0))
rd2 <- ilm_rdd(covs, "score", "run", cutoff = 0,
               covariates = c("prior", "leaky"), verbose = FALSE)
rd2$balance
```

`leaky` jumps, which means something other than treatment changes at the cutoff
and the estimate absorbs it.

**Placebo cutoffs and bandwidth sensitivity.**

```{r}
rd$placebo
rd$bandwidth
```

An effect that appears only in a narrow window is not an effect.

### What this does not do

The interval is the ordinary one for a weighted local fit. It does **not** carry
the bias correction that a wide bandwidth calls for; `rdrobust` implements the
Calonico–Cattaneo–Titiunik intervals and does it well. The default here is local
**linear** for a reason: a high-order global polynomial produces estimates driven
by points far from the cutoff and is a documented way to manufacture a
discontinuity.

Fuzzy assignment -- where crossing the cutoff changes the *probability* of
treatment without determining it -- is detected and not reported as sharp, since
the jump is then the effect of being eligible rather than of being treated.

## Choosing among the three

| | identifying assumption | checkable? |
|---|---|---|
| `ilm_dag_model()` | the graph is right, and the adjustment set is measured | partly: the implied independencies |
| `ilm_did()` | parallel trends | partly: the pre-treatment periods |
| `ilm_rdd()` | nothing else changes at the cutoff | partly: density, balance, placebos |

None of them is checkable where it matters, which is the part that concerns a
counterfactual. Everything above makes an assumption more or less plausible; none
of it makes one true.

## See also

`vignette("regression-models")` for the modelling engine and its diagnostics,
`vignette("exploring-data", package = "illumex")` for the exploration side, and
`vignette("effect-size-and-power")` for `ilm_scenario()`, which turns a fitted
causal model into the projection a decision needs. Mediation -- how much of an
effect runs THROUGH something else -- is `ilm_mediate()`, documented in
`vignette("regression-models")`.
