---
title: "Effect size and power"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect size and power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Three questions that bracket an analysis. Before it: how large does the study
need to be. After it: how large is the effect, on a scale someone can act on.
And then: what would happen if we changed something.

```{r}
library(illume)
```

## Before: how large a study

`ilm_power()` treats a fitted model as the truth, simulates studies at the sizes
you are considering, refits each one, and counts how often the term is detected.
Nothing in it needs a closed-form variance, so it covers mixed models,
zero-inflated counts and ordinal outcomes as readily as a linear regression.

```{r}
pilot <- ilm_model(y ~ dose, data = pilot_data, family = "binomial")
pw <- ilm_power(pilot, n = c(200, 400, 800), effect = c(0.3, 0.5), sims = 500)
pw
plot(pw)
```

Two things it reports that most software does not, and both change what the
number means.

### The power estimate is itself an estimate

A power of 0.80 from 200 replicates is a proportion, so it carries a standard
error of `sqrt(0.8 * 0.2 / 200)`, about 2.8%. The power is 0.74 to 0.86, and a
sample size read off that curve is a range rather than a number.

```{r}
ilm_power_n(pw, target = 0.8)
#>   effect     n  n_lower  n_upper
#>      0.3   240      207      272
```

Quoting "you need 240 participants" from that is false precision. Quoting
"somewhere between 210 and 270, and more replicates will narrow it" is the
answer the simulation supports. Raising `sims` narrows the interval and nothing
else does -- collecting more pilot data changes the estimate, not its Monte
Carlo error.

The interval is Wilson rather than Wald, because power estimates live near 1,
where a Wald interval reaches above it.

### Some fits will not converge

A replicate that fails to fit has not detected anything. `power` counts it as a
non-detection; `power_converged` divides only by the replicates that worked,
which is **power conditional on convergence** -- a different and more
flattering quantity, and the one most software reports.

```{r}
#>   n effect power mc_lower mc_upper power_converged converged
#>  40  2.072 0.850    0.756    0.912           0.872     0.975
```

When `converged` is below one the two differ, and the gap is the size of the
problem rather than something to smooth over. A design where a fifth of studies
fail to fit is a design worth reconsidering, and averaging over only the
successes hides exactly that.

### Mixed designs

For a model with a grouping factor, `n` is the number of rows but the number of
**clusters** moves with it, because that is what a mixed design's power actually
depends on. Resampling rows would hold the number of groups fixed while n grew,
which is the wrong thing to hold fixed. A cluster drawn twice becomes two
clusters, or the design has fewer independent groups than it appears to.

## After: how large the effect

`ilm_effects()` reports each coefficient on the scale its family puts it on.

```{r}
fit <- ilm_model(relapse ~ dose + age, data = d, family = "binomial")
ilm_effects(fit)
```

| family | scale |
|---|---|
| binomial, beta | odds ratio |
| poisson, negative binomial | incidence rate ratio |
| ordinal | proportional odds ratio |
| weibull, lognormal, loglogistic | time ratio (an acceleration factor, **not** a hazard ratio) |
| Royston-Parmar | hazard ratio |
| gaussian | the coefficient, plus partial variance explained |

Three details that are easy to get wrong.

**The interval is built on the link scale and transformed.** `exp()` of the
endpoints is exact and asymmetric, which is the shape a ratio's uncertainty
really has. A standard error computed on the ratio scale and doubled either way
is symmetric and can reach below zero.

**A ratio from a mixed model is conditional.** `exp(beta)` there is the odds
ratio for a *given cluster* -- two patients in the same hospital, one exposed --
not for the population, and the gap widens as the random effects grow. The
print says so and names `ilm_ame()` for the population-averaged quantity:

```{r}
ilm_effects(fit)      # conditional odds ratio, e.g. 2.58
ilm_ame(fit, "dose")  # marginal effect on the probability scale, e.g. 0.193
```

**Standardising a binary predictor is meaningless.** Dividing by the standard
deviation of a 0/1 variable answers what happens per standard deviation of
treatment, which is not a question anyone asks. Those are left alone, and the
reason is printed rather than the number quietly omitted. `standardise =
"gelman"` divides numeric predictors by *two* standard deviations, which is what
makes them comparable with a binary one.

```{r}
ilm_effects(fit, standardise = "gelman")
```

A robust covariance can be substituted, so the same table can be built on
clustered or design-based standard errors:

```{r}
ilm_effects(fit, vcov = ilm_vcov_cluster(fit, ~ clinic))
```

## Then: what would happen if

`ilm_scenario()` sets predictors to values someone is considering and reports
what the model implies.

```{r}
ilm_scenario(fit, dose = c(0, 10, 20), contrast = "first")
```

The default averages over **the observed units**: set the intervention for
everyone, keep each unit's own covariates, predict, average. That is
standardisation, and when the adjustment set is valid it is the causal
estimand.

This is not the same as building one row with the predictors set and predicting
from it. That row is a *representative unit* -- what happens to someone who
looks like this -- and under any non-linear link it is a different number,
because the average of a prediction is not the prediction at the average. On a
logistic fit at the same dose: 0.657 averaged over the units, 0.683 at the
average covariate. `over = "reference"` gives the second, and the print says
which was used.

```{r}
ilm_scenario(fit, dose = 10)                        # the population
ilm_scenario(fit, dose = 10, over = "reference")    # a unit like this
```

### It checks whether the scenario is inside the data

A model returns a number for a dose nobody received, and the interval around it
looks like any other, because the interval carries uncertainty about the
coefficients rather than about whether the functional form survives out there.
Each scenario is checked twice.

**By value**, against each predictor's observed range -- with a 2% tolerance, so
asking for `dose = 0` when the smallest observed is 0.013 is not flagged.
Flagging that every time is how a warning gets ignored.

**By combination**, which no range check catches. Every value can sit inside its
own range while the point sits in a corner nobody occupies. On personnel data
where years of service is roughly age minus 22, a 25-year-old with 35 years of
service has both values well inside their ranges, and sits 2.39 standardised
units from the nearest real person against 0.25 for a typical one.

```{r}
ilm_scenario(fit, age = c(45, 25), service = c(23, 35))
#>  age service estimate lower upper extrapolating
#>   45      23    3.095 3.003 3.180         FALSE
#>   25      35    2.820 1.292 4.192          TRUE
```

A flagged scenario is reported rather than refused. Sometimes extrapolating is
the point -- but the reader should know they are doing it, and nothing in the
number tells them.

## See also

`vignette("workflow")` for where these sit in an analysis, `?ilm_power`,
`?ilm_effects`, `?ilm_scenario`, and `?ilm_ame` for the marginal effect of a
one-unit change rather than a named scenario.
