---
title: "Effect size and power"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect size and power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Three questions that bracket an analysis. Before it: how large does the study
need to be. After it: how large is the effect, on a scale someone can act on.
And then: what would happen if we changed something.

```{r}
library(illume)
```

## Before: how large a study

`ilm_power()` treats a fitted model as the truth, simulates studies at the sizes
you are considering, refits each one, and counts how often the term is detected.
Nothing in it needs a closed-form variance, so it covers mixed models,
zero-inflated counts and ordinal outcomes as readily as a linear regression.

```{r}
pilot <- ilm_model(y ~ dose, data = pilot_data, family = "binomial")
pw <- ilm_power(pilot, n = c(200, 400, 800), effect = c(0.3, 0.5), sims = 500)
pw
plot(pw)
```

Two things it reports that most software does not, and both change what the
number means.

### The power estimate is itself an estimate

A power of 0.80 from 200 replicates is a proportion, so it carries a standard
error of `sqrt(0.8 * 0.2 / 200)`, about 2.8%. The power is 0.74 to 0.86, and a
sample size read off that curve is a range rather than a number.

```{r}
ilm_power_n(pw, target = 0.8)
#>   effect     n  n_lower  n_upper
#>      0.3   240      207      272
```

Quoting "you need 240 participants" from that is false precision. Quoting
"somewhere between 210 and 270, and more replicates will narrow it" is the
answer the simulation supports. Raising `sims` narrows the interval and nothing
else does -- collecting more pilot data changes the estimate, not its Monte
Carlo error.

The interval is Wilson rather than Wald, because power estimates live near 1,
where a Wald interval reaches above it.

### Some fits will not converge

A replicate that fails to fit has not detected anything. `power` counts it as a
non-detection; `power_converged` divides only by the replicates that worked,
which is **power conditional on convergence** -- a different and more
flattering quantity, and the one most software reports.

```{r}
#>   n effect power mc_lower mc_upper power_converged converged
#>  40  2.072 0.850    0.756    0.912           0.872     0.975
```

When `converged` is below one the two differ, and the gap is the size of the
problem rather than something to smooth over. A design where a fifth of studies
fail to fit is a design worth reconsidering, and averaging over only the
successes hides exactly that.

### The test it counts is the one you will report

Each simulated study is analysed the way the real one will be. A model with
nothing to integrate out reports exact t and F tests, so that is what is
counted; anything else reports a Wald z, and a term with several coefficients
-- a factor with three levels, or any term of a multinomial model, which has
one coefficient per category -- is tested jointly, as `ilm_anova()` tests it.

```{r}
ilm_power(pilot, term = "dose")       # one coefficient: its z or t test
ilm_power(pilot3, term = "arm")       # three arms: the joint test of both
ilm_power(pilot3, term = "armB")      # or one comparison, named as a coefficient
```

It matters most where studies are small. Counting a z test where the analysis
will use a t put the power of a 20-person two-arm trial at 0.43, against an
exact 0.395.

For a joint test `effect` is a **multiple** of the assumed coefficients rather
than a value for one of them: `effect = c(0.5, 1)` asks what happens if every
part of the effect is half what was assumed.

A study is also refitted as the same model -- the contrasts, the zero part,
the dispersion model, the censoring and a flexible survival baseline all come
along, and serial correlation is simulated and refitted as itself. Each of
those was once dropped, and a model with sum-to-zero contrasts came back with
differently named coefficients and a power of exactly zero.

### Mixed designs

For a model with a grouping factor, `n` is the number of rows but the number of
**clusters** moves with it, because that is what a mixed design's power actually
depends on. Resampling rows would hold the number of groups fixed while n grew,
which is the wrong thing to hold fixed. A cluster drawn twice becomes two
clusters, or the design has fewer independent groups than it appears to --
and so do the classes inside a school drawn twice, while items that every
participant saw keep their labels, because new participants do not bring new
items.

A random **slope** is drawn as a random slope, not as an intercept shift. A
`(1 + time | id)` design whose participants were all given the same slope would
have them moving in near-parallel, which understates how far they differ and so
overstates power for anything interacting with `time`.

### When there is no pilot study

`ilm_power()` treats a fitted model as the truth, and a study being planned has
nothing to fit. `ilm_scaffold()` builds the model from assumptions instead: a
design grid, the parameters you are prepared to defend, and an ordinary
`ilm_model` object that everything else in the package will accept.

Assumptions go in as cell means on the response scale, which is the form most
people actually hold one in.

```{r}
d <- list(arm = c("control", "treatment"), time = c("pre", "post"))
cells <- c(control.pre = 20, control.post = 21,
           treatment.pre = 20, treatment.post = 25)

s <- ilm_scaffold(change ~ arm * time + (1 | id), design = d,
                  within = "time", n_unit = 60,
                  cells = cells, sd = 6, icc = 0.5)
#> <ilm_scaffold>  parameters ASSUMED, not estimated
#>   formula    : change ~ arm * time + (1 | id)
#>   family     : gaussian
#>   size       : 60 ids x 2 = 120 rows
#>   between    : arm
#>   within     : time
#>
#>   assumed coefficients
#>           (Intercept)          armtreatment              timepost
#>                    20                     0                     1
#> armtreatment:timepost
#>                     4
#>
#>   residual sd : 6
#>   id sd       : 6
#>   total sd    : 8.485   <- the outcome's own spread; `sd` above is the residual
```

`n_unit` counts participants, not rows, so the number you give is the number
that goes in a protocol. `arm` varies between participants and `time` within,
and the grid is built accordingly -- allocating a between-participants variable
to rows would put the same person in both arms.

The point of returning a model rather than a number is that the assumptions can
be interrogated before anything is built on them. The cell means went in; here
they are coming back out through an entirely different route.

```{r}
ilm_emmeans(s, c("arm", "time"))
#>        arm time estimate    se lower upper
#>    control  pre       20 1.595 16.87 23.13
#>  treatment  pre       20 1.595 16.87 23.13
#>    control post       21 1.595 17.87 24.13
#>  treatment post       25 1.595 21.87 28.13
```

Cell means the formula cannot produce are refused rather than approximated. A
least-squares solve always returns *something*, and the something it returns for
crossed means under an additive formula is a scaffold for a different study:

```{r}
ilm_scaffold(change ~ arm + time, design = d, within = "time",
             n_unit = 60, cells = cells, sd = 6, icc = 0.5)
#> Error: this formula cannot produce the cell means given: cell
#> 'control.pre' is off by 1 on the link scale. Cell means that differ by
#> more than the terms in the formula allow need the interaction between
#> them -- arm * time rather than arm + time.
```

`ilm_power_design()` is the scaffold plus the power curve in one call.

```{r}
pw <- ilm_power_design(change ~ arm * time + (1 | id), design = d,
                       within = "time", n_unit = c(60, 120, 240, 400),
                       cells = cells, sd = 6, icc = 0.5,
                       term = "arm:time", sims = 300)
#> Simulated power for armtreatment:timepost (gaussian family, alpha = 0.05)
#>   counting the Wald z test, as the analysis reports it
#>   300 replicates per cell; the scaffold's own grid had 800 rows
#>   each replicate is a fresh draw of the planned design: balanced
#>   allocation, new participants
#>  n_unit   n effect power mc_lower mc_upper power_converged converged
#>      60 120      4 0.490    0.434    0.546           0.490         1
#>     120 240      4 0.747    0.695    0.793           0.747         1
#>     240 480      4 0.957    0.927    0.975           0.957         1
#>     400 800      4 0.983    0.962    0.993           0.983         1

ilm_power_n(pw, target = 0.8)
#>  effect   n n_lower n_upper n_unit n_unit_lower n_unit_upper
#>       4 301   249.8   348.7  150.5        124.9        174.4
```

Read `n_unit`, not `n`: about 150 participants measured twice -- somewhere
between 125 and 175, given the Monte Carlo error -- not 300 people.
`ilm_interpret(pw)` writes the same thing out in sentences.

The standard errors printed on a scaffold come from **one realisation** of the
design at its own size. They are not a property of the assumptions, and power
read off them would be one coin flip. That is what `ilm_power_design()` is for.

None of this makes an assumption true. It makes the assumption explicit, and
separates "we need 150 participants" from the four numbers that claim implies.

### `sd` is the residual standard deviation

Worth stating plainly, because every other package that takes a design
specification means something different by the word. `sd` is the **residual**
scale, and the random effects sit on top of it, so with `icc` the outcome's own
spread is `sd / sqrt(1 - icc)`. The example above — `sd = 6, icc = 0.5` — is
data whose standard deviation is 8.49, not 6.

`Superpower` and `faux` both take the **total** standard deviation, and so does
a paper you might read one off. Converting is:

```
  sd_illume = sd_total * sqrt(1 - icc)
```

The scaffold prints both so the two cannot be quietly confused:

```{r}
#>   residual sd : 4.243
#>   id sd       : 4.243
#>   total sd    : 6   <- the outcome's own spread; `sd` above is the residual
```

With a random slope the total varies by row — the spread depends on where you
are in the design — and it says that instead of printing a number true at one
point only.

### Every simulated study is a fresh draw of the plan

A scaffold has no data to resample, so each replicate draws the planned design
again at its own size: participants allocated to cells the way a protocol
randomises a factorial -- every combination of the between-participant
factors equally often, and any remainder spread so each factor's own levels
stay balanced -- and a covariate given as a function drawn afresh. Resampling
the scaffold's own grid instead, as this once did, left cells of a small
factorial empty: 18 participants in a 3 x 3 design fitted 28.5% of the time,
where the planned study, with two in every cell, always can.

### A categorical outcome

For a multinomial or ordinal outcome the assumption is the probability of each
category in each cell, which is how it would be written in a protocol.

```{r}
cells <- rbind(control   = c(none = 0.50, some = 0.30, full = 0.20),
               treatment = c(none = 0.35, some = 0.35, full = 0.30))
pw <- ilm_power_design(y ~ arm, design = list(arm = c("control", "treatment")),
                       n_unit = c(100, 200, 400), family = "multinomial",
                       cells = cells, term = "arm", sims = 1000)
#>  n_unit   n effect power mc_lower mc_upper power_converged converged
#>     100 100      1 0.239    0.214    0.266           0.239         1
#>     200 200      1 0.468    0.437    0.499           0.468         1
#>     400 400      1 0.818    0.793    0.841           0.818         1
```

`arm` is tested across all its categories at once, and `effect` is a multiple
of the assumed coefficients. Checked against the noncentral chi-square the
asymptotic theory gives, the curve is 0.818 against 0.813 at 400; at 100 and
200 it sits 3 to 4 points below theory (0.239 against 0.272, 0.468 against
0.502), because the Wald test the analysis reports is itself a little
conservative there -- it rejected a true null 4.2% of the time at 100 and 4.5%
at 200. A Pearson test of the 2 x 3 table would have more power at those sizes,
and that is not the test a mixed-model analysis reports.

An ordinal outcome can be stated the same way, or as slopes and `thresholds`.
Probabilities no proportional-odds model can produce are refused, with the
model that can produce them named:

```{r}
bad <- rbind(control = c(lo = 0.38, mid = 0.35, hi = 0.27),
             treatment = c(lo = 0.20, mid = 0.60, hi = 0.20))
ilm_scaffold(y ~ arm, design = list(arm = c("control", "treatment")),
             n_unit = 150, family = "ordinal", cells = bad)
#> Error: these cell probabilities are not ones proportional odds can produce:
#> between cells, the cumulative log-odds shift by different amounts at
#> different cut points (off by 0.322). An ordinal model would plan a
#> different study from the one described. family = "multinomial" reproduces
#> these probabilities exactly.
```

A random intercept in a multinomial model is a random shift in every
category's log-odds, and `re_sd` is its standard deviation. For a binary or
ordinal outcome an `icc` works as it does for a gaussian one, on the latent
scale, whose residual variance is that of the standard logistic, `pi^2 / 3`.

### When the design itself fails a check

Some checks the analysis makes are made *before* fitting, on the design, so
every study drawn from it fails them alike. Counting that against each
replicate would put the power at zero whatever the effect, so the result
reports it instead -- with the remedy the check itself names:

```{r}
#>   The DESIGN fails a check the analysis makes before fitting, and
#>   every analysis of such a study will report it as FAILED:
#>     latent_budget: 2.00 observations per latent value (240 observations,
#>       120 latent values: id 120)  (100% of the simulated studies)
#>       try: ... term 'id' contributes 120 (us, width 2). rr(1) would cut it to 60
```

That one is a three-category multinomial with four visits per participant: a
random intercept per category and participant leaves two observations for
every value it has to estimate. The same design with eight visits does not
fail it.

With 60 participants the Wald test of the treatment effect in that design
rejected a true null 5.8% of the time with four visits and 7.8% with eight, so
the power reported is a little flattered, by the same amount. That is the test
the analysis will report; `ilm_pb_lrt()` is the calibrated alternative for the
analysis itself.

### What it was checked against

Simulated power is only as good as the simulation, so this was compared with
three outside implementations on designs each could express.

**`power.t.test`**, two-sample, effect 2.5 on sd 4, 2,000 replicates:

```
  participants     100      200      300
  illume         0.878    0.995    1.000
  closed form    0.872    0.993    1.000
```

And where it matters most, at 20 participants with an effect of 0.8 standard
deviations: 0.387 (0.374 to 0.399 over 6,000 replicates) against the exact
0.395. Counting a z test there, as this once did, gave 0.43.

**`Superpower::ANOVA_exact()`** and **`simr::makeLmer()`** on the 2 (arm,
between) × 2 (time, within) design above, cell means 20/21/20/25 with `sd = 6`
and an ICC of 0.5. `simr` is the closest existing thing to a scaffold — it
builds a model from `fixef`, `VarCorr` and `sigma` with nothing fitted — so it
is the fairest comparison available. `simr` 1,000 replicates, illume 2,000:

```
  n per arm   Superpower   simr (z)   simr (Satterthwaite)   illume (ML)   illume (REML)
     20          0.306       0.329           0.302              0.358         0.342
     40          0.549       0.552           0.545              0.583         0.572
```

**`faux`**, which generates the data a completely different way, on a random
slope design. Per-subject moments from 8,000 simulated participants:

```
                intercept sd   slope sd   correlation
  target            2.000        1.200       -0.400
  faux              2.157        1.270       -0.432
  illume            2.156        1.273       -0.446
```

Checked directly as well: the random effects drawn for 8,000 simulated groups
have standard deviations 2.016 and 1.209 and a correlation of −0.413, against
2, 1.2 and −0.4 assumed.

Two things worth taking from that. The agreement with `faux` to three decimals
is the strongest evidence that random slopes are simulated correctly — they
were not, until recently, and an independent generator is what settled it.

And illume sits above Superpower's exact calculation, for two reasons that can
be taken apart. It counts the test its analysis reports, and a mixed model's
default analysis is maximum likelihood with a z reference. Maximum likelihood's
variance components run small, which flatters the power by about 1.5 points
here; `reml = TRUE` plans for a REML analysis instead, and brings illume
within Monte Carlo error of `simr`, which uses REML too. What remains is z
against an exact F: a Satterthwaite reference takes `simr` down to
Superpower's value. Expect gaps of that size when comparing against G\*Power or
Superpower, and choose `reml` to match the analysis you will actually run.

## After: how large the effect

`ilm_effects()` reports each coefficient on the scale its family puts it on.

```{r}
fit <- ilm_model(relapse ~ dose + age, data = d, family = "binomial")
ilm_effects(fit)
```

| family | scale |
|---|---|
| binomial, beta | odds ratio |
| poisson, negative binomial | incidence rate ratio |
| ordinal | proportional odds ratio |
| weibull, lognormal, loglogistic | time ratio (an acceleration factor, **not** a hazard ratio) |
| Royston-Parmar | hazard ratio |
| gaussian | the coefficient, plus partial variance explained |

Three details that are easy to get wrong.

**The interval is built on the link scale and transformed.** `exp()` of the
endpoints is exact and asymmetric, which is the shape a ratio's uncertainty
really has. A standard error computed on the ratio scale and doubled either way
is symmetric and can reach below zero.

**A ratio from a mixed model is conditional.** `exp(beta)` there is the odds
ratio for a *given cluster* -- two patients in the same hospital, one exposed --
not for the population, and the gap widens as the random effects grow. The
print says so and names `ilm_ame()` for the population-averaged quantity:

```{r}
ilm_effects(fit)      # conditional odds ratio, e.g. 2.58
ilm_ame(fit, "dose")  # marginal effect on the probability scale, e.g. 0.193
```

**Standardising a binary predictor is meaningless.** Dividing by the standard
deviation of a 0/1 variable answers what happens per standard deviation of
treatment, which is not a question anyone asks. Those are left alone, and the
reason is printed rather than the number quietly omitted. `standardise =
"gelman"` divides numeric predictors by *two* standard deviations, which is what
makes them comparable with a binary one.

```{r}
ilm_effects(fit, standardise = "gelman")
```

A robust covariance can be substituted, so the same table can be built on
clustered or design-based standard errors:

```{r}
ilm_effects(fit, vcov = ilm_vcov_cluster(fit, ~ clinic))
```

## Then: what would happen if

`ilm_scenario()` sets predictors to values someone is considering and reports
what the model implies.

```{r}
ilm_scenario(fit, dose = c(0, 10, 20), contrast = "first")
```

The default averages over **the observed units**: set the intervention for
everyone, keep each unit's own covariates, predict, average. That is
standardisation, and when the adjustment set is valid it is the causal
estimand.

This is not the same as building one row with the predictors set and predicting
from it. That row is a *representative unit* -- what happens to someone who
looks like this -- and under any non-linear link it is a different number,
because the average of a prediction is not the prediction at the average. On a
logistic fit at the same dose: 0.657 averaged over the units, 0.683 at the
average covariate. `over = "reference"` gives the second, and the print says
which was used.

```{r}
ilm_scenario(fit, dose = 10)                        # the population
ilm_scenario(fit, dose = 10, over = "reference")    # a unit like this
```

### It checks whether the scenario is inside the data

A model returns a number for a dose nobody received, and the interval around it
looks like any other, because the interval carries uncertainty about the
coefficients rather than about whether the functional form survives out there.
Each scenario is checked twice.

**By value**, against each predictor's observed range -- with a 2% tolerance, so
asking for `dose = 0` when the smallest observed is 0.013 is not flagged.
Flagging that every time is how a warning gets ignored.

**By combination**, which no range check catches. Every value can sit inside its
own range while the point sits in a corner nobody occupies. On personnel data
where years of service is roughly age minus 22, a 25-year-old with 35 years of
service has both values well inside their ranges, and sits 2.39 standardised
units from the nearest real person against 0.25 for a typical one.

```{r}
ilm_scenario(fit, age = c(45, 25), service = c(23, 35))
#>  age service estimate lower upper extrapolating
#>   45      23    3.095 3.003 3.180         FALSE
#>   25      35    2.820 1.292 4.192          TRUE
```

A flagged scenario is reported rather than refused. Sometimes extrapolating is
the point -- but the reader should know they are doing it, and nothing in the
number tells them.

## See also

`vignette("workflow")` for where these sit in an analysis, `?ilm_power`,
`?ilm_scaffold` and `?ilm_power_design` for planning without a pilot,
`?ilm_effects`, `?ilm_scenario`, and `?ilm_ame` for the marginal effect of a
one-unit change rather than a named scenario.
