---
title: "Exploring data before you model it"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring data before you model it}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
set.seed(2026)
library(illume)
```

## What this half of the package is for

Most exploratory tools answer "what does this data look like". `illume` asks a
narrower question: **what about this data will break a model?**

That is a different emphasis, and it shows up everywhere in the output. A
factor summary reports how many levels are too thin to estimate, because that
is the usual reason a model fails to fit. A date summary reports whether the
spacing is regular, because an AR(1) term requires it. A numeric summary
reports how far the variable is from gaussian *and why*, because the reason
determines what you do next.

The functions share one prefix with the modelling half, so exploring and
fitting feel like one workflow rather than two packages bolted together.

```{r}
d <- ilm_sim()
dim(d)
```

`ilm_sim()` generates a synthetic dataset for examples and tests. Every column
is there because some check should have something to say about it — a fixture
where nothing is wrong tests nothing.

## Describing everything at once

```{r}
r <- ilm_describe_all(d)
names(r)
```

One table per class, because columns of different classes do not share a
sensible set of statistics. Constant columns are split out separately: a
variable with one value tells you nothing a statistic can show, and it will
break a model matrix.

```{r}
r$numeric
```

Reading across: `p_zero` flags zero inflation, `dispersion` is the
variance-to-mean ratio for count-like variables, and `gauss` with `gauss_note`
say how far the variable is from normal and why.

```{r}
r$constant
```

## The gaussian index, and why it is not a p-value

`gauss` runs from 0 to 1. A value of 1 means the departure from normality is no
larger than sampling noise would produce at this sample size.

```{r}
ilm_gauss_check(rnorm(500))$gauss
ilm_gauss_check(rlnorm(500))$gauss
```

It deliberately is **not** a normality test p-value. Any test of exact
normality rejects everything once *n* is large, so the p-value ends up
answering a question nobody asked — you already know real data is not exactly
normal. What matters is *how far* from normal it is, which is an effect size.

```{r}
# a normality test would reject this; the index does not
ilm_gauss_check(rnorm(1e5))$gauss
```

The measure underneath is the Kolmogorov distance to the best-fitting normal:
the largest amount, on the cumulative probability scale, by which a normal
model misstates the data. Ask for it directly with `gauss = "ks_d"` if you
would rather apply your own threshold.

```{r}
ilm_describe(d, "income", gauss = "both")[, c("gauss", "ks_d", "gauss_note")]
```

### The reason matters more than the number

```{r}
notes <- function(x) ilm_gauss_check(x)$gauss_note
notes(rlnorm(2000))
notes(rt(2000, 3))
notes(rpois(2000, 3))
notes(c(rnorm(1000), rnorm(1000, 5)))
```

That last one is the case worth dwelling on. A mixture of two normals is
**multimodal**, and the advice — check for subgroups — is what you actually
need. An earlier version of this feature tried to name the single best-fitting
distribution family instead, and on that data it confidently answered
"uniform". It was not close to right, and it offered nothing to act on.

## What the summaries are looking for

```{r}
r$categorical[, c("variable", "n_empty", "n_unique", "n_rare", "n_unused",
                  "case_variants", "note")]
```

Each of those columns exists because of a specific failure:

- `n_empty` — `""` is not `NA` to R, but it is almost always missing to you,
  and it silently becomes a factor level.
- `n_rare` — levels with only a handful of observations are the usual reason a
  factor model will not fit, and in a logistic model a direct route to
  separation.
- `n_unused` — a declared but unobserved level produces an all-zero column in
  the model matrix and then breaks `predict()` on new data.
- `case_variants` — `"North"`, `"north"` and `"North "` counted as three
  things.

```{r}
r$logical[, c("variable", "p_TRUE", "note")]
```

An indicator that is almost always `TRUE` is a separation risk in any binary
model.

### Dates carry the AR(1) precondition

```{r}
r$time[, c("variable", "n_unique", "spacing", "regular", "n_gaps", "note")]
```

`regular` and `n_gaps` matter because `ilm_model()`'s AR(1) term needs
regularly spaced observations within group. Seeing "irregular, 3 gaps" here is
much better than discovering it when the model fails.

Repeated dates are *described* rather than warned about: many units sharing a
date is simply what long-format panel data looks like.

## Problems that belong to pairs of columns

Some problems are not properties of any single variable.

```{r}
dd <- d
dd$score_copy <- dd$score
ilm_frame_issues(dd)
```

Note that `score` is not reported as identifier-like even though every value is
distinct — a continuous variable is unique per row by construction. Only
discrete-valued columns earn that flag.

## Missingness

```{r}
head(ilm_describe_na_all(d), 4)
```

Sorted with the most missing first, since those are the variables worth
looking at.

## Plots that carry the verdict

`ilm_plot()` chooses a plot from the classes of what you give it, and annotates
it with the same verdict `ilm_describe()` reports.

```{r}
ilm_plot(d, "income")
```

The subtitle is not decoration. A histogram of a skewed variable looks like a
histogram; the annotation says how far from gaussian it is and in which
direction.

You can ask what would be drawn without drawing it:

```{r}
ilm_pick_geom(d$score)$geom
ilm_pick_geom(d$grp, d$score)$geom
```

### Rendering is aware of how much data there is

```{r}
big <- data.frame(a = rnorm(50000), b = rnorm(50000))
ilm_pick_geom(big$a, big$b)
```

A scatter of fifty thousand points is a black rectangle. Above `n_max` the plot
becomes a binned density, and says so rather than misleading you quietly.

```{r}
ilm_plot(big, "a", "b")
```

### Choosing the geom yourself

```{r}
ilm_geom_spec()
```

That table is not just documentation — it is what the argument checks read, so
the documented requirements and the enforced ones cannot drift apart. Ask for
something impossible and the error says what would work:

```{r, error = TRUE, purl = FALSE}
ilm_plot(d, "grp", geom = "histogram")
```

### Appearance

Colours, fills, transparency, point size, palettes and themes are all
available, with ggplot2-style names:

```{r}
ilm_plot(d, "grp", "score", geom = "violin", fill = "lightblue",
         alpha = 0.7, theme = "clean")
```

```{r}
ilm_plot(d, "score", "income", by = "grp", palette = "Dark 2", alpha = 0.6)
```

## Cleaning

```{r}
messy <- data.frame(
  "Study ID"  = c("1", "2", "3", ""),
  someFlag    = c("TRUE", "FALSE", "TRUE", ""),
  measurement = c("1.5", "2.5", "n/a", ""),
  allEmpty    = c("", "", "", ""),
  check.names = FALSE, stringsAsFactors = FALSE)

ilm_wash_df(messy)
```

Names become snake_case, the empty row and column go, and text columns that are
really numeric or logical are converted. `measurement` stays character on
purpose: one value does not convert, and silently turning a column into `NA`s
is worse than leaving it alone.

Known-bad codes are handled separately, because a sentinel that means missing
in one column can be a real measurement in another:

```{r}
ilm_recode_errors(data.frame(x = c(1, 999, 3), y = c(999, 2, 3)),
                  errors = 999, cols = "x")
```

And recoding against a lookup table:

```{r}
ilm_translate(c(1, 2, 3, 99), old = 1:3, new = c("low", "mid", "high"))
```

Unmatched values become `NA` rather than passing through, so codes your
dictionary does not cover cannot hide.

## Duplicates and counts

```{r}
ilm_counts(d$grp)
ilm_counts_tb(d$site, n = 2)
```

`ilm_counts_tb()` shows both ends at once because that is where the problems
are: a level that dominates, and levels too thin to model.

```{r}
dup <- data.frame(a = c(1, 1, 2, 2, 2, 3), b = c("x", "x", "y", "y", "z", "w"))
ilm_copies(dup)
```

## Uncertainty without a model

```{r}
ilm_boot_ci(d, "score", R = 500, seed = 1)
```

For a difference between two groups, ask for the difference directly:

```{r}
d2 <- d[d$grp %in% c("alpha", "beta"), ]
d2$grp <- factor(d2$grp)
ilm_boot_diff(d2, "score", "grp", R = 500, seed = 1)
```

`excludes_zero` answers the question people actually ask. Comparing two
separate intervals for overlap is a conservative and lossy substitute: two
intervals can overlap while the difference is clearly non-zero.

Past two groups every pair is compared, and a formula works in place of the
two column names:

```{r}
ilm_boot_diff(score ~ grp, data = d, R = 500, seed = 1)
```

Six comparisons at a nominal 95% each do not jointly cover at 95%, so the
intervals are **simultaneous** by default: the groups are resampled together,
each comparison is standardised by its own bootstrap standard error, and the
largest standardised value in each replicate supplies one critical value for
all of them. On normal, equal-variance data this reproduces `TukeyHSD()` to
about 0.006, without assuming either normality or a common variance. Set
`adjust = "none"` for comparisons chosen in advance, or `adjust =
"bonferroni"` to keep the shape of a percentile interval.

`ref` compares every level against one:

```{r}
ilm_boot_diff(score ~ grp, data = d, ref = "alpha", R = 500, seed = 1)
```

On skewed data, `"percentile"` and `"bca"` hold their nominal coverage better
than `"normal"` and `"basic"`. When the skew is severe and the groups are
small, no interval shape rescues a bootstrapped *mean* -- reach for
`stat = "median"` instead, which stays nominal where the mean does not.

## Unusual values

```{r outliers}
head(ilm_outliers_all(mtcars), 5)
```

Three rules, in decreasing order of robustness. `"iqr"` is the Tukey fence and
reproduces exactly what a boxplot's whiskers draw — it uses Tukey's hinges, not
`quantile()`'s default, which matters on small samples and is pinned to
`grDevices::boxplot.stats()` by a test. `"mad"` measures from the median in
median absolute deviations. `"zscore"` measures from the mean in standard
deviations, and is the weakest of the three for a reason worth seeing:

```{r outliers2}
z <- c(rnorm(30), 100)
c(mad = max(ilm_outliers(z, "mad")$score),
  zscore = max(ilm_outliers(z, "zscore")$score))
```

One large value inflates the standard deviation enough to hide itself. The
median and the MAD are not dragged around by the values being detected.

Grouping changes the answer, and usually should:

```{r outliers3}
nrow(ilm_outliers_all(mtcars))
nrow(ilm_outliers_all(mtcars, by = "cyl"))
```

A value can be perfectly ordinary for its own group and extreme against the
pooled distribution, so flagging without `by` on grouped data partly just
rediscovers the groups.

**A flag is not a verdict.** It says a value is far from the others by some
rule — which may mean a typo, a different population, or an ordinary draw from
a heavy tail, and the third is common. Dropping flagged rows because they are
flagged changes what you are estimating. If the tail is real, the fix is a model
that expects it: a different family, or `ilm_model(dispformula = )`.

## Naming the plot you want

`ilm_plot()` chooses a geometry from the data. When you would rather say which
one, each has its own function — `ilm_plot_histogram()`, `ilm_plot_density()`,
`ilm_plot_box()`, `ilm_plot_violin()`, `ilm_plot_scatter()`, `ilm_plot_bar()`,
`ilm_plot_line()`, `ilm_plot_stat_error()` — and they all take column names as
strings, like everything else here.

```{r namedplots, fig.height = 3.2}
ilm_plot_c(
  ilm_plot_histogram(mtcars, "mpg"),
  ilm_plot_box(mtcars, "mpg", x = "cyl"),
  nrow = 1
)
```

`ilm_plot_var_pairs()` handles every pair at once, mixed column types included,
and `ilm_plot_var_all()` draws one panel per column.

The bootstrap difference from earlier can be seen rather than only summarised:

```{r bootplot, fig.height = 3.6}
b <- ilm_boot_diff(d, "score", "grp", R = 500, seed = 1)
ilm_plot_boot_diff(b, row = 1)
```

The interval says where the difference is. This says what the resampling
produced — whether it is symmetric, skewed, or piled against a boundary, which
the two endpoints cannot show.

## Profiling: what shape is this data set?

`ilm_describe_all()` tells you about columns one at a time. `ilm_profile()`
asks a different question -- are there *kinds* of row here? -- by reducing the
columns to a few dimensions, clustering on those, and then saying what each
cluster is.

```{r profile, eval = requireNamespace("PCAmixdata", quietly = TRUE) && requireNamespace("cluster", quietly = TRUE)}
cars <- mtcars
cars$cyl <- factor(cars$cyl)
cars$am  <- factor(cars$am)

p <- ilm_profile(cars, k_max = 5, B = 25, seed = 1)
cat(p$summary, sep = "\n\n")
```

The reduction picks its own method from the column types: PCA when they are all
numeric, multiple correspondence analysis when they are all categorical, and a
mixed method when both are present, as here.

Two things get reported that a bare cluster number does not give you.
**Stability** is a property of the cluster -- resample the rows, recluster, and
see whether the same grouping comes back. **Ambiguity** is a property of the
row, from its silhouette width, and is independent of cluster size: a point can
sit on the boundary between two clusters inside a large and perfectly stable
one, which a size-based check would never surface.

```{r profilebits, eval = requireNamespace("PCAmixdata", quietly = TRUE) && requireNamespace("cluster", quietly = TRUE)}
p$cluster$clusters
```

A cluster is characterised by v-test, the same device `FactoMineR`'s `catdes()`
uses. It is a threshold rather than a test, and is documented as one: the
clusters were found from the very coordinates being tested, so the usual
sampling argument does not apply. What speaks to whether a partition is real is
the stability and silhouette figures above, not the characterisation.

## Missing values

`ilm_describe_na_all()` counts them. The harder question is whether they matter,
and that has a more useful answer than it usually gets.

```{r missing}
set.seed(4)
n  <- 500
z  <- rnorm(n)
x  <- 0.6 * z + rnorm(n)
y  <- 0.5 * x + 0.3 * z + rnorm(n)
d2 <- data.frame(x = x, z = z, y = y)
d2$x[plogis(1.2 * z - 0.4) > runif(n)] <- NA   # missingness follows z

ilm_check_missing(d2, y = "y", verbose = FALSE)
```

Note the verdict. Missingness here depends on `z` — so this is emphatically not
MCAR — and yet **complete-case analysis is unbiased**. Dropping incomplete rows
needs missingness to be independent of the *outcome* given the covariates, which
is far weaker than MCAR and is satisfied here. Imputing would buy back some
precision and nothing else.

That distinction is why the check asks the question conditionally. Missingness
driven by `z` is *marginally* associated with `y`, because `y` depends on `z`
too; reading the verdict off that marginal association would send you off to
impute for nothing.

Change what drives the missingness and the advice changes with it:

```{r missingout}
d3 <- d2
d3$x <- x
d3$x[plogis(1.0 * y - 0.4) > runif(n)] <- NA   # now it follows the outcome
ilm_check_missing(d3, y = "y", verbose = FALSE)$verdict
```

Measured over 200 replicates at about 40% missing, the coefficient on `x` came
back with a bias of 0.0001 and 98% interval coverage in the first case, against
a bias of −0.099 — a fifth of the effect — and 61.5% coverage in the second.

### Imputing, when it is actually needed

```{r impute}
imp <- ilm_impute(d3, m = 10, seed = 1, verbose = FALSE)
ilm_mi_pool(imp, y ~ x + z, family = "gaussian")
```

Each missing value is **drawn** from the predictive distribution rather than set
to a fitted mean, and the `m` completed data sets are pooled by Rubin's rules,
where the spread of the estimates *across* imputations becomes part of the
reported uncertainty. `fmi` is the fraction of information lost to missingness.

`single = TRUE` gives one completed data set and warns. Filling values in once
and analysing as though they had been observed gets the point estimate about
right and the standard errors badly wrong: in that same simulation, single
imputation covered 79% to 89% where the nominal rate was 95%, because nothing
in its standard errors knows part of the data was invented.

### Which values go missing together

```{r profilena, eval = requireNamespace("PCAmixdata", quietly = TRUE) && requireNamespace("cluster", quietly = TRUE)}
pn <- ilm_profile_na(airquality, k_max = 4, B = 25, seed = 1)
cat(pn$summary[1:2], sep = "\n\n")
```

The same reduce-cluster-describe pipeline, pointed at present/missing indicators
instead of values. A block of columns that go missing as one points at a shared
cause — a section of a form a whole group skipped — which is a different problem
from values going one at a time, and calls for a different fix.

And the boundary of all this is worth stating plainly: **MAR and MNAR cannot be
told apart from the observed data.** That is a theorem, not a gap in the
implementation — the data that would separate them are the ones that are
missing. Everything above tests MCAR and describes what missingness is related
to among the things you can see. If the chance of a value going missing depends
on the value itself, no test will reveal it, and the remedy is a sensitivity
analysis rather than a better diagnostic.

## Into the model

The point of all this is what comes next. The exploration tells you which
family to consider, which variables will cause trouble, and whether an AR(1)
term is even possible:

```{r}
f <- ilm_model(score ~ income + grp + (1 | id), data = d,
               family = "gaussian", verbose = FALSE)
ilm_plot_model(f, "coef")
```

and, once it is fitted, what it says in words:

```{r}
ilm_interpret(f, ame = FALSE)
```

See `vignette("illume")` for the modelling half, and
`vignette("causal-models")` for turning an association into a causal claim.
