---
title: "Introduction to illume"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to illume}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2024)
```

## What this package is for

`illume` is a frequentist regression workflow aimed at **inference** rather than
prediction. One function, `ilm_model()`, fits everything from a simple
regression to a mixed model with smooths and correlated errors, across five
response distributions. Because a single engine fits them all, the object, the
methods and the diagnostics do not change as the model gets more complicated.

```r
ilm_model(y ~ x + (1 | subj), data = dd, family = "gaussian")
ilm_model(y ~ x + (1 | subj), data = dd, family = "poisson")
ilm_model(y ~ x + (1 | subj), data = dd, family = "multinomial")
```

The **multinomial** family closes a specific gap. If your outcome has three or
more categories with no natural order -- a diagnosis, a habitat type, a choice
among options -- and your data are grouped, R has had no good frequentist
option: `nnet::multinom()` has no random effects, `lme4::glmer()` and `glmmTMB`
have no multinomial family, and `brms` is much slower and answers a different
question.

Most of this vignette uses the multinomial family because it is the most
involved case. Everything shown applies to the other families unchanged.

```{r, eval = FALSE}
library(illume)
```

```{r, include = FALSE}
library(illume)
```

## A first model

We will simulate data where an outcome with three categories depends on a
continuous predictor `x1` and a grouping factor `grp`, measured repeatedly within
subjects.

```{r simulate}
n_subj <- 40; per <- 20; N <- n_subj * per
dd <- data.frame(
  subj = factor(rep(seq_len(n_subj), each = per)),
  x1   = rnorm(N),
  grp  = factor(sample(c("a", "b", "c"), N, TRUE))
)

# true structure: an effect of x1 and grp, plus subject-level variation
X  <- model.matrix(~ x1 + grp, dd)
bt <- matrix(c(0.1, -0.2,  0.9, 0.5,  -0.6, 0.4,  0.2, -0.5), 4, 2, byrow = TRUE)
b  <- matrix(rnorm(n_subj * 2), n_subj, 2) %*% diag(c(0.8, 0.6))
eta <- X %*% bt + b[as.integer(dd$subj), ]
P <- exp(eta %*% t(contr.sum(3))); P <- P / rowSums(P)
dd$y <- factor(c("low", "mid", "high")[apply(P, 1, function(p) sample.int(3, 1, prob = p))],
               levels = c("low", "mid", "high"))

table(dd$y)
```

The model is written the way you would write it in `lme4`:

```{r fit}
fit <- ilm_model(y ~ x1 + grp + (1 | subj), data = dd,
                 family = "multinomial", verbose = FALSE)
fit
```

## Reading the output

```{r summary}
summary(fit)
```

Two things in that output deserve attention.

**The coefficients are sum-to-zero contrasts.** A coefficient such as `low:x1` is
the effect of `x1` on the `low` category *relative to the average across all
categories*, not relative to a baseline category. This differs from
`nnet::multinom()` and `brms`, which compare each category against a reference.
The summary prints a reminder, because it is easy to misread if you are used to
the other convention.

**The assumption checks are printed with the coefficients.** That is unusual for
a `summary()` method and it is deliberate, for a reason worth explaining.

## Why the diagnostics are so prominent

This class of model fails *quietly*. A fit can converge, produce entirely
reasonable-looking coefficients with small standard errors and significance
stars, and yet the covariance matrix behind those standard errors can be
unusable -- meaning the uncertainty is meaningless even though nothing in the
output looks wrong.

Every fit therefore carries a table of checks:

```{r checks}
head(fit$checks[, c("check", "status", "detail")], 8)
```

Each check has one of five statuses. `OK` and `FAIL` are self-explanatory. `WARN`
means something is worth a look. `BOUNDARY` means a variance sits at zero, where
the usual tests do not apply. And `INCONCLUSIVE` means there is not enough
information to judge -- which is reported honestly rather than being passed off
as a clean bill of health.

Anything that is not `OK` also carries a reason and a suggested remedy:

```{r check-detail}
bad <- fit$checks[fit$checks$status != "OK", ]
if (nrow(bad)) bad[1, c("check", "cause", "suggestion")] else "all checks passed"
```

### The most useful check

The one to look at first is `latent_budget`. A mixed model introduces one
unobserved value per random effect per group. The method used to fit the model
(the Laplace approximation) is accurate when each of those values is informed by
a reasonable amount of data, and degrades when it is not.

Importantly this is a *global* constraint. A single term can look perfectly
healthy on its own and the model can still be over-stretched, because what
matters is the total. Per-term checks cannot see that; this one can.

## Testing fixed effects

`ilm_anova()` gives an analysis of deviance. Each term is tested **jointly
across all category dimensions**, so a predictor with one column is tested on two
degrees of freedom when there are three categories.

```{r anova}
ilm_anova(fit, type = 3)
```

Type III tests each term with all others present; Type II tests each term after
the terms that do not contain it. They agree when there are no interactions. Wald
tests are the default because they need no refitting; a likelihood-ratio version
is available with `test = "LRT"` and is more reliable for very strong effects.

For small samples, neither is ideal -- both rely on large-sample approximations,
and unlike linear mixed models there is no Kenward-Roger correction available
here. `ilm_pb_lrt()` replaces the assumed reference distribution with a
simulated one:

```{r pb, eval = FALSE}
ilm_pb_lrt(fit, "x1", B = 200, ncores = 4)
```

## Predictions

By default `predict()` returns a probability for every observation and category,
as `nnet::multinom(type = "probs")` does.

```{r predict}
head(round(predict(fit), 3))
```

There is a choice to make here. With `marginal = FALSE` (the default) the random
effects are set to zero, giving probabilities for a *typical* subject. With
`marginal = TRUE` the prediction is averaged over the distribution of subjects,
giving probabilities for the *population*.

```{r marginal}
grid <- data.frame(x1 = 0, grp = factor(c("a", "b", "c"), levels = levels(dd$grp)))
round(predict(fit, newdata = grid, marginal = FALSE), 3)
round(predict(fit, newdata = grid, marginal = TRUE, ndraw = 200), 3)
```

These differ because averaging and the softmax transform do not commute: the
average of the transformed values is not the transform of the average. The
population-averaged probabilities are pulled toward being more even across
categories. Which you want depends on the question you are asking.

Standard errors and intervals come from simulation, and the intervals are
percentile intervals, so they always lie between 0 and 1.

```{r se}
pr <- predict(fit, newdata = grid, se.fit = TRUE, nsim = 100)
round(data.frame(fit = pr$fit[, 1], lower = pr$lower[, 1], upper = pr$upper[, 1]), 3)
```

## Random slopes, smooths and correlated time

Beyond random intercepts, the formula accepts random slopes and penalised
smooths:

```{r more, eval = FALSE}
ilm_model(y ~ x1 + time + (1 + time | subj), data = dd, family = "multinomial")
ilm_model(y ~ x1 + s(xs, k = 10) + (1 | subj), data = dd, family = "multinomial")
ilm_model(y ~ x1 + t2(lon, lat) + (1 | site), data = dd, family = "multinomial")
```

Two rules for smooths: use `t2()` rather than `te()` for tensor products, and
write them unqualified -- `s(x)`, not `mgcv::s(x)` -- because mgcv identifies
smooths by name.

## Simple models, and exact inference

Nothing requires a random effect. `ilm_model()` fits an ordinary regression, and
when the family is gaussian and nothing is integrated out it reports **exact**
inference rather than large-sample approximations:

```{r lmlike}
set.seed(1)
sd2 <- data.frame(x = rnorm(200), z = factor(sample(c("a", "b", "c"), 200, TRUE)))
sd2$v <- 2 + 1.5 * sd2$x - 0.8 * (sd2$z == "b") + rnorm(200, 0, 1.2)

f_lm <- ilm_model(v ~ x + z, data = sd2, family = "gaussian", verbose = FALSE)
round(ilm_coef_table(f_lm), 5)
round(summary(lm(v ~ x + z, data = sd2))$coefficients, 5)
```

The two agree to numerical precision, including the residual standard deviation
and the degrees of freedom. `ilm_anova()` likewise switches from chi-square to F:

```{r lmanova}
ilm_anova(f_lm, type = 3)
```

This matters because the chi-square test treats the residual variance as known,
which it is not, making it anti-conservative in small samples. Where an exact
reference exists, illume uses it.

It applies only where it is genuinely available. A Poisson or binomial model
without random effects is still a GLM, with no exact small-sample analogue, so
it keeps z and chi-square. So does any model with a random effect.

## Other families

The same formula syntax, methods and diagnostics apply to every family. What
changes is the response, the link, and whether there is a dispersion parameter:

```{r families, eval = FALSE}
ilm_model(count ~ x + (1 | site), data = dd, family = "poisson")
ilm_model(count ~ x + (1 | site), data = dd, family = "nbinom")   # overdispersed
ilm_model(hit   ~ x + (1 | site), data = dd, family = "binomial")
ilm_model(score ~ x + (1 | site), data = dd, family = "gaussian")
```

`summary()` reports a dispersion parameter where the family has one -- the
residual standard deviation for gaussian, the overdispersion parameter for
negative binomial -- and omits that section otherwise. Coefficients carry no
category prefix outside the multinomial, and the sum-to-zero note disappears,
because there is only one linear predictor.

A two-category outcome can be stored however your data happen to hold it --
`0`/`1`, a two-level factor, `TRUE`/`FALSE`, or a character column read straight
from a csv. All four give the same fit. For a factor or character response the
*second* level is the one being modelled, exactly as in `glm()`, so
`factor(c("no", "yes"))` models the probability of `"yes"`. If you want the other
direction, relevel the factor. You do not have to remember which way round it is:
`summary()` states the category it is modelling, and the reference it is measured
against.

Two categories belong to `"binomial"` and three or more to `"multinomial"`;
each family says so if it is given the other one's data.

If your counts vary more than Poisson allows, `"nbinom"` adds one parameter `k`,
giving variance `mu + mu^2/k`. Smaller `k` means more overdispersion; large `k`
approaches the Poisson.

Zero-inflation, hurdle models and Tweedie are deliberately out of scope.
`glmmTMB` covers those well, and a weaker reimplementation would help nobody.

## When there are many categories

The number of covariance parameters grows quickly with the number of categories.
With an unstructured covariance and `J` categories you need `(J-1)J/2` parameters
*per grouping factor* -- 3 when `J = 3`, but 45 when `J = 10`. Many designs
cannot support that.

The remedy is a reduced-rank structure, which describes the covariance with fewer
dimensions:

```{r rr, eval = FALSE}
ilm_model(y ~ x1 + (1 | subj), data = dd, family = "multinomial",
          re_struct = list(subj = list(type = "rr", rank = 3)))
```

This saves parameters *and* latent values, and the latter is usually the bigger
gain. When a structure is too rich for the data, the checks say so and name a
specific rank to try.

## Checking the fitted model

`ilm_appraise()` draws diagnostic panels built for a categorical outcome rather
than borrowed from tools designed for continuous responses:

```{r appraise, fig.width = 7, fig.height = 5, eval = FALSE}
ilm_appraise(fit)
```

Two further checks are worth knowing about, because broad residual summaries have
very little power to detect the problems they address. `ilm_check_omitted()`
tests variables you did *not* include, and `ilm_check_ar()` tests for leftover
correlation over time:

```{r targeted, eval = FALSE}
ilm_check_omitted(fit, dd)
ilm_check_ar(fit, time = dd$time, group = dd$subj)
```

And `ilm_consistency()` asks the most direct question available: can the model
recover itself from data it generated? If not, the fitting method is not reliable
for this model and this amount of data.

```{r consistency, eval = FALSE}
ilm_consistency(fit, B = 50, ncores = 4)
```

## Working with other packages

`illume` fits into the usual ecosystem. `car::Anova()` dispatches to the correct
joint test, `performance::model_performance()` and `check_model()` work, and
`marginaleffects` can be enabled with one call:

```{r ecosystem, eval = FALSE}
ilm_register_marginaleffects()
marginaleffects::avg_slopes(fit, variables = "x1")
```

## Where to read more

The model itself is a baseline-category logit with random effects; Agresti
(2013), *Categorical Data Analysis*, chapter 8, is the standard reference. The
fitting method is described in Kristensen et al. (2016), *Journal of Statistical
Software* 70(5). The mixed-model representation of smooths follows Wood (2017),
*Generalized Additive Models*, section 5.4. The parametric bootstrap follows
Halekoh and Højsgaard (2014), *Journal of Statistical Software* 59(9).
