---
title: "Introduction to illume"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to illume}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2024)
library(illume)
```

## What this package is for

`illume` is a frequentist analysis workflow aimed at **inference** rather than
prediction. It covers the path from a sample-size calculation, through
exploration and missing data, to a model, its diagnostics, and a scenario a
stakeholder can act on -- with one object and one vocabulary the whole way.

Two commitments shape it.

**One engine fits everything.** `ilm_model()` goes from a simple regression to a
mixed model with smooths and correlated errors, across families from gaussian
through counts, proportions, ordered and unordered categories and survival
times. Because a single engine fits them all, `summary()`, `ilm_anova()`,
`predict()` and every diagnostic stay the same as the model gets harder.

**Every check names a remedy that exists in this package.** A diagnostic that
tells you an assumption fails and leaves you to find the fix elsewhere is a
complaint. `ilm_check_variance()` names `dispformula`; `ilm_check_zeros()` names
`ziformula`; `ilm_check_proportional()` names `family = "multinomial"`. Where a
remedy did not exist, it was built -- which is most of what this package is.

```{r, eval = FALSE}
library(illume)
```

## A worked example, end to end

```{r endtoend, eval = FALSE}
d <- ilm_sim()                       # a grouped dataset with a known structure

## what is in it
ilm_describe_all(d)
ilm_check_missing(d, downtime ~ income + region)

## a model: a count outcome, repeated within worker
fit <- ilm_model(downtime ~ income + region + (1 | id), data = d,
                 family = "poisson")

## does it hold up?
ilm_appraise(fit)                    # the whole battery at once
ilm_check_zeros(fit)                 # more zeros than poisson allows?

## what does it say?
ilm_anova(fit)
ilm_effects(fit)                     # incidence rate ratios
em <- ilm_emmeans(fit, "region")
ilm_contrast(em)                     # every pair, simultaneously

## what would happen if we changed something?
ilm_scenario(fit, income = c(20, 40, 60), contrast = "first")

## say it in words
ilm_interpret(fit)
```

If `ilm_check_zeros()` flags excess zeros, the next line is
`ilm_model(..., ziformula = ~ 1)` and everything downstream is unchanged. That
is the shape of the whole package.

## What is in it

| | |
|---|---|
| **Models** | `ilm_model()` -- gaussian, binomial, poisson, negative binomial, beta, multinomial, three ordinal links, three accelerated failure time families, Royston-Parmar survival; random effects, smooths, AR(1)/CAR(1), dispersion models, zero-inflation and hurdles |
| **Other designs** | `ilm_iv()` instrumental variables, `ilm_did()` difference in differences, `ilm_rdd()` regression discontinuity, `ilm_design()` complex samples |
| **Diagnostics** | `ilm_appraise()` and around twenty individual checks, each naming its remedy |
| **Inference** | `ilm_anova()`, `ilm_effects()`, `ilm_emmeans()`/`ilm_contrast()`, `ilm_ame()`, `ilm_robust()`, `ilm_pb_lrt()`, `ilm_boot_ci()` |
| **Exploration** | `ilm_describe_all()`, fifteen `ilm_plot_*()` functions, `ilm_outliers()`, `ilm_anomaly()` |
| **Structure** | `ilm_reduce()`, `ilm_cluster()`, `ilm_profile()`, `ilm_glrm()` |
| **Missing data** | `ilm_check_missing()`, `ilm_impute()`, `ilm_mi_pool()` |
| **Causal** | `ilm_dag()`, `ilm_adjust_sets()`, `ilm_dag_test()`, `ilm_dag_model()`, `ilm_mediate()` |
| **Design and decision** | `ilm_power()`, `ilm_scenario()`, `ilm_interpret()` |

## The multinomial claim, stated carefully

If an outcome has three or more categories with no natural order -- a
diagnosis, a habitat type, a choice among options -- and the data are grouped,
the frequentist options are thin. `nnet::multinom()` has no random effects;
`lme4::glmer()` and `glmmTMB` have no multinomial family; `brms` is Bayesian
and much slower.

```{r, eval = FALSE}
ilm_model(choice ~ price + (1 | household), data = dd, family = "multinomial")
```

**They are not absent, though.** `mclogit::mblogit()` fits this model, by
penalised quasi-likelihood rather than by maximising a Laplace-approximated
marginal likelihood. The difference is measurable, and it is not uniform:

```
                       illume    mclogit
  40 clusters x 25
    coverage              0.953     0.949      <- no meaningful difference
    attenuation           1.005     0.960
  100 clusters x 4
    coverage              0.949     0.891
    attenuation           1.004     0.731      <- 27% shrunk toward zero
    bias                  0.035     0.188
    RMSE                  0.409     0.336      <- mclogit WINS here
```

With well-populated clusters the two agree and there is nothing to choose
between them. PQL degrades where the clusters are small, because it shrinks the
fixed effects toward zero -- and its standard errors do not know it, so the
interval is centred in the wrong place at close to the right width.

The last row is the honest caveat: mclogit's **RMSE is lower**, because
shrinkage buys a variance reduction that more than pays for the bias. If the
goal is prediction that is a defensible trade. If the goal is a coefficient
with an interval you intend to interpret, it is not, and coverage is the
measure that says so. `studies/findings/mclogit.md` has the full table.

## Working with other packages

`illume` fits into the usual ecosystem. `car::Anova()` dispatches to the correct
joint test, `performance::model_performance()` and `check_model()` work, and
`marginaleffects` and the easystats stack can each be enabled with one call:

```{r ecosystem, eval = FALSE}
ilm_register_marginaleffects()
marginaleffects::avg_slopes(fit, variables = "x1")

ilm_register_insight()          # parameters, performance, report
parameters::model_parameters(fit)
```

`ilm_ame()` agrees with `marginaleffects::avg_slopes()` to 1e-6 on the estimates
and 1e-7 on the standard errors, which is the point of having both: an
independent implementation is the only thing that has ever found a real bug
here.

## Where to read more

`vignette("workflow")` is the map: eleven stages from power to reporting, with a
pointer for each.

| Vignette | Covers |
|---|---|
| `workflow` | the whole path, stage by stage |
| `regression-models` | `ilm_model()` in depth, every family, mixed multinomial |
| `exploring-data` | descriptives, plots, bootstrap intervals |
| `profiling` | `ilm_reduce()`, `ilm_cluster()`, `ilm_profile()` |
| `anomaly-detection` | rows that are implausible as combinations |
| `missing-data` | diagnosing it, imputing it, pooling |
| `causal-models` | DAGs, difference in differences, discontinuities |
| `effect-size-and-power` | effect sizes, power, scenario projection |

## References

The multinomial model is a baseline-category logit with random effects; Agresti
(2013), *Categorical Data Analysis*, chapter 8, is the standard reference. The
fitting method is described in Kristensen et al. (2016), *Journal of Statistical
Software* 70(5). The mixed-model representation of smooths follows Wood (2017),
*Generalized Additive Models*, section 5.4. The parametric bootstrap follows
Halekoh and Højsgaard (2014), *Journal of Statistical Software* 59(9).
