---
title: "Introduction to illume"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to illume}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2024)
```

## What this package is for

`illume` is a frequentist regression workflow aimed at **inference** rather than
prediction. One function, `ilm_model()`, fits everything from a simple
regression to a mixed model with smooths and correlated errors, across five
response distributions. Because a single engine fits them all, the object, the
methods and the diagnostics do not change as the model gets more complicated.

```r
ilm_model(y ~ x + (1 | subj), data = dd, family = "gaussian")
ilm_model(y ~ x + (1 | subj), data = dd, family = "poisson")
ilm_model(y ~ x + (1 | subj), data = dd, family = "multinomial")
```

The **multinomial** family closes a specific gap. If your outcome has three or
more categories with no natural order -- a diagnosis, a habitat type, a choice
among options -- and your data are grouped, R has had no good frequentist
option: `nnet::multinom()` has no random effects, `lme4::glmer()` and `glmmTMB`
have no multinomial family, and `brms` is much slower and answers a different
question.

Most of this vignette uses the multinomial family because it is the most
involved case. Everything shown applies to the other families unchanged.

```{r, eval = FALSE}
library(illume)
```

```{r, include = FALSE}
library(illume)
```

## A first model

We will simulate data where an outcome with three categories depends on a
continuous predictor `x1` and a grouping factor `grp`, measured repeatedly within
subjects.

```{r simulate}
n_subj <- 40; per <- 20; N <- n_subj * per
dd <- data.frame(
  subj = factor(rep(seq_len(n_subj), each = per)),
  x1   = rnorm(N),
  grp  = factor(sample(c("a", "b", "c"), N, TRUE))
)

# true structure: an effect of x1 and grp, plus subject-level variation
X  <- model.matrix(~ x1 + grp, dd)
bt <- matrix(c(0.1, -0.2,  0.9, 0.5,  -0.6, 0.4,  0.2, -0.5), 4, 2, byrow = TRUE)
b  <- matrix(rnorm(n_subj * 2), n_subj, 2) %*% diag(c(0.8, 0.6))
eta <- X %*% bt + b[as.integer(dd$subj), ]
P <- exp(eta %*% t(contr.sum(3))); P <- P / rowSums(P)
dd$y <- factor(c("low", "mid", "high")[apply(P, 1, function(p) sample.int(3, 1, prob = p))],
               levels = c("low", "mid", "high"))

table(dd$y)
```

The model is written the way you would write it in `lme4`:

```{r fit}
fit <- ilm_model(y ~ x1 + grp + (1 | subj), data = dd,
                 family = "multinomial", verbose = FALSE)
fit
```

## Reading the output

```{r summary}
summary(fit)
```

Two things in that output deserve attention.

**The coefficients are sum-to-zero contrasts.** A coefficient such as `low:x1` is
the effect of `x1` on the `low` category *relative to the average across all
categories*, not relative to a baseline category. This differs from
`nnet::multinom()` and `brms`, which compare each category against a reference.
The summary prints a reminder, because it is easy to misread if you are used to
the other convention.

**The assumption checks are printed with the coefficients.** That is unusual for
a `summary()` method and it is deliberate, for a reason worth explaining.

## Why the diagnostics are so prominent

This class of model fails *quietly*. A fit can converge, produce entirely
reasonable-looking coefficients with small standard errors and significance
stars, and yet the covariance matrix behind those standard errors can be
unusable -- meaning the uncertainty is meaningless even though nothing in the
output looks wrong.

Every fit therefore carries a table of checks:

```{r checks}
head(fit$checks[, c("check", "status", "detail")], 8)
```

Each check has one of five statuses. `OK` and `FAIL` are self-explanatory. `WARN`
means something is worth a look. `BOUNDARY` means a variance sits at zero, where
the usual tests do not apply. And `INCONCLUSIVE` means there is not enough
information to judge -- which is reported honestly rather than being passed off
as a clean bill of health.

Anything that is not `OK` also carries a reason and a suggested remedy:

```{r check-detail}
bad <- fit$checks[fit$checks$status != "OK", ]
if (nrow(bad)) bad[1, c("check", "cause", "suggestion")] else "all checks passed"
```

### The most useful check

The one to look at first is `latent_budget`. A mixed model introduces one
unobserved value per random effect per group. The method used to fit the model
(the Laplace approximation) is accurate when each of those values is informed by
a reasonable amount of data, and degrades when it is not.

Importantly this is a *global* constraint. A single term can look perfectly
healthy on its own and the model can still be over-stretched, because what
matters is the total. Per-term checks cannot see that; this one can.

## Testing fixed effects

`ilm_anova()` gives an analysis of deviance. Each term is tested **jointly
across all category dimensions**, so a predictor with one column is tested on two
degrees of freedom when there are three categories.

```{r anova}
ilm_anova(fit, type = 3)
```

Type III tests each term with all others present; Type II tests each term after
the terms that do not contain it. They agree when there are no interactions. Wald
tests are the default because they need no refitting; a likelihood-ratio version
is available with `test = "LRT"` and is more reliable for very strong effects.

For small samples, neither is ideal -- both rely on large-sample approximations,
and unlike linear mixed models there is no Kenward-Roger correction available
here. `ilm_pb_lrt()` replaces the assumed reference distribution with a
simulated one:

```{r pb, eval = FALSE}
ilm_pb_lrt(fit, "x1", B = 200, ncores = 4)
```

## Predictions

By default `predict()` returns a probability for every observation and category,
as `nnet::multinom(type = "probs")` does.

```{r predict}
head(round(predict(fit), 3))
```

There is a choice to make here. With `marginal = FALSE` (the default) the random
effects are set to zero, giving probabilities for a *typical* subject. With
`marginal = TRUE` the prediction is averaged over the distribution of subjects,
giving probabilities for the *population*.

```{r marginal}
grid <- data.frame(x1 = 0, grp = factor(c("a", "b", "c"), levels = levels(dd$grp)))
round(predict(fit, newdata = grid, marginal = FALSE), 3)
round(predict(fit, newdata = grid, marginal = TRUE, ndraw = 200), 3)
```

These differ because averaging and the softmax transform do not commute: the
average of the transformed values is not the transform of the average. The
population-averaged probabilities are pulled toward being more even across
categories. Which you want depends on the question you are asking.

Standard errors and intervals come from simulation, and the intervals are
percentile intervals, so they always lie between 0 and 1.

```{r se}
pr <- predict(fit, newdata = grid, se.fit = TRUE, nsim = 100)
round(data.frame(fit = pr$fit[, 1], lower = pr$lower[, 1], upper = pr$upper[, 1]), 3)
```

## Random slopes, smooths and correlated time

Beyond random intercepts, the formula accepts random slopes and penalised
smooths:

```{r more, eval = FALSE}
ilm_model(y ~ x1 + time + (1 + time | subj), data = dd, family = "multinomial")
ilm_model(y ~ x1 + s(xs, k = 10) + (1 | subj), data = dd, family = "multinomial")
ilm_model(y ~ x1 + t2(lon, lat) + (1 | site), data = dd, family = "multinomial")
```

Two rules for smooths: use `t2()` rather than `te()` for tensor products, and
write them unqualified -- `s(x)`, not `mgcv::s(x)` -- because mgcv identifies
smooths by name.

## Simple models, and exact inference

Nothing requires a random effect. `ilm_model()` fits an ordinary regression, and
when the family is gaussian and nothing is integrated out it reports **exact**
inference rather than large-sample approximations:

```{r lmlike}
set.seed(1)
sd2 <- data.frame(x = rnorm(200), z = factor(sample(c("a", "b", "c"), 200, TRUE)))
sd2$v <- 2 + 1.5 * sd2$x - 0.8 * (sd2$z == "b") + rnorm(200, 0, 1.2)

f_lm <- ilm_model(v ~ x + z, data = sd2, family = "gaussian", verbose = FALSE)
round(ilm_coef_table(f_lm), 5)
round(summary(lm(v ~ x + z, data = sd2))$coefficients, 5)
```

The two agree to numerical precision, including the residual standard deviation
and the degrees of freedom. `ilm_anova()` likewise switches from chi-square to F:

```{r lmanova}
ilm_anova(f_lm, type = 3)
```

This matters because the chi-square test treats the residual variance as known,
which it is not, making it anti-conservative in small samples. Where an exact
reference exists, illume uses it.

It applies only where it is genuinely available. A Poisson or binomial model
without random effects is still a GLM, with no exact small-sample analogue, so
it keeps z and chi-square. So does any model with a random effect.

## Other families

The same formula syntax, methods and diagnostics apply to every family. What
changes is the response, the link, and whether there is a dispersion parameter:

```{r families, eval = FALSE}
ilm_model(count ~ x + (1 | site), data = dd, family = "poisson")
ilm_model(count ~ x + (1 | site), data = dd, family = "nbinom")   # overdispersed
ilm_model(hit   ~ x + (1 | site), data = dd, family = "binomial")
ilm_model(score ~ x + (1 | site), data = dd, family = "gaussian")
```

`summary()` reports a dispersion parameter where the family has one -- the
residual standard deviation for gaussian, the overdispersion parameter for
negative binomial -- and omits that section otherwise. Coefficients carry no
category prefix outside the multinomial, and the sum-to-zero note disappears,
because there is only one linear predictor.

A two-category outcome can be stored however your data happen to hold it --
`0`/`1`, a two-level factor, `TRUE`/`FALSE`, or a character column read straight
from a csv. All four give the same fit. For a factor or character response the
*second* level is the one being modelled, exactly as in `glm()`, so
`factor(c("no", "yes"))` models the probability of `"yes"`. If you want the other
direction, relevel the factor. You do not have to remember which way round it is:
`summary()` states the category it is modelling, and the reference it is measured
against.

Two categories belong to `"binomial"` and three or more to `"multinomial"`;
each family says so if it is given the other one's data.

If your counts vary more than Poisson allows, `"nbinom"` adds one parameter `k`,
giving variance `mu + mu^2/k`. Smaller `k` means more overdispersion; large `k`
approaches the Poisson.

Zero-inflation, hurdle models and Tweedie are deliberately out of scope.
`glmmTMB` covers those well, and a weaker reimplementation would help nobody.

## When there are many categories

The number of covariance parameters grows quickly with the number of categories.
With an unstructured covariance and `J` categories you need `(J-1)J/2` parameters
*per grouping factor* -- 3 when `J = 3`, but 45 when `J = 10`. Many designs
cannot support that.

The remedy is a reduced-rank structure, which describes the covariance with fewer
dimensions:

```{r rr, eval = FALSE}
ilm_model(y ~ x1 + (1 | subj), data = dd, family = "multinomial",
          re_struct = list(subj = list(type = "rr", rank = 3)))
```

This saves parameters *and* latent values, and the latter is usually the bigger
gain. When a structure is too rich for the data, the checks say so and name a
specific rank to try.

## Checking the fitted model

`ilm_appraise()` draws diagnostic panels built for a categorical outcome rather
than borrowed from tools designed for continuous responses:

```{r appraise, fig.width = 7, fig.height = 5, eval = FALSE}
ilm_appraise(fit)
```

Two further checks are worth knowing about, because broad residual summaries have
very little power to detect the problems they address. `ilm_check_omitted()`
tests variables you did *not* include, and `ilm_check_ar()` tests for leftover
correlation over time:

```{r targeted, eval = FALSE}
ilm_check_omitted(fit, dd)
ilm_check_ar(fit, time = dd$time, group = dd$subj)
```

And `ilm_consistency()` asks the most direct question available: can the model
recover itself from data it generated? If not, the fitting method is not reliable
for this model and this amount of data.

```{r consistency, eval = FALSE}
ilm_consistency(fit, B = 50, ncores = 4)
```

## Structure over time

Repeated measurements on the same unit are rarely independent, and there are two
quite different reasons why -- which matter because they need opposite remedies.

`ilm_plot_acf()` draws the residual autocorrelation and partial autocorrelation
by lag, within group, with a verdict on each lag:

```{r acfplot, eval = FALSE}
ilm_plot_acf(fit, time = dd$time, group = dd$subj, maxlag = 14)
```

Three things make this not `stats::acf()`:

- Pairs are matched on exact time differences **within group**. A panel dataset
  stacked long is many short series, and pairing across the joins manufactures
  correlation that is not there.
- The band is not `2/sqrt(n)`. Residuals within a group sum to roughly zero, so
  pairs of them correlate *negatively* even when the model is exactly right --
  measured, around -0.07 at every lag on a 60-unit panel. A band centred on zero
  calls correct models wrong. The band here comes from simulating from the fit
  and refitting.
- The band is drawn at the critical value the p-value is computed against, so a
  point outside the band is exactly a lag the check flags. The picture and the
  table cannot disagree.

### When the times are not evenly spaced

`ilm_ar1()` needs a common step, because "one step" has to mean something.
When visits fall where they fall, use `ilm_car1()`: the correlation between two
observations is `rho` raised to the gap between them, which is `nlme`'s
`corCAR1()` and reduces to AR(1) when every gap is one.

```{r car1, eval = FALSE}
ilm_model(y ~ x + (1 | subj), data = dd, family = "gaussian",
          ar = ilm_car1(dd$day, dd$subj))
```

Both constructors report their **latent budget** before you fit, because that is
the constraint that bites. These are latent-variable structures: one latent
value exists per distinct unit-and-time, so genuinely continuous times give one
latent per observation. For a gaussian response that is fine — the model is
linear-Gaussian, the Laplace approximation is exact, and coverage measured over
400 replicates at exactly one observation per latent came back at 0.955, 0.937
and 0.949 against a nominal 0.95. For a categorical response it is not, and the
fit says so. Rounding `time` to a coarser grid, so observations share a latent,
is the usual fix.

### Checking a correlation structure over irregular time

`ilm_check_ar()` matches pairs at exact lags, which stops working once the times
are irregular: on a sixty-unit panel with six observations each drawn from
thirty possible times, it found 53 pairs at lag 1, five at lag 2 and none
beyond. `ilm_variogram()` bins every within-group pair by how far apart it is
instead, and uses all 900.

```{r variogram, eval = FALSE}
ilm_variogram(fit, time = dd$day, group = dd$subj)
```

It distinguishes three shapes, because they need three different remedies:

| what you see | what it is | what to do |
|---|---|---|
| decays from the shortest separation | autoregression | `ilm_car1()` |
| the same level at every separation | a group effect | add or widen a random effect |
| alternates in sign | a cycle | `ilm_fourier()`, `ilm_cyclic()` |

The envelope again comes from refitting simulated data, and again is not centred
on zero: residuals within a group sum to roughly zero, so pairs of them
correlate negatively even under a correct model, and more so the further apart
they are.

### Correlation in space

The same function takes coordinates instead of a time, because separation in
space and separation in time are the same statistic over a different distance:

```{r spatialvg, eval = FALSE}
ilm_variogram(fit, coords = dd[, c("easting", "northing")])
```

The remedy it names is different, and deliberately so. illume fits no spatial
covariance — no Matern field, no exponential one — so the advice does not
pretend otherwise. What it points at is putting the structure in the **mean**:
a tensor-product smooth of the coordinates, `t2(x, y)`, absorbs smooth spatial
variation, and a random effect for a spatial grouping absorbs the coarse kind.

Measured on 300 points carrying an exponential field, residual correlation ran
0.242 at the shortest distance down through 0.071 to −0.072 at the longest.
Adding `t2(sx, sy)` cut the first of those to −0.033, dropped AIC from 1165.7 to
1069.4, and narrowed the standard error on the covariate of interest from 0.100
to 0.083.

### A departure too small to act on

Each bin rests on thousands of pairs, so the envelope narrows until any
imperfection clears it. In the example above the residual correlation after the
smooth was about 0.03 against an envelope 0.02 wide — a seventh of what it had
been, and nothing anyone would act on — and four of five bins were still
flagged.

`min_effect` is the floor, 0.1 by default: a bin outside the envelope by less
than that is reported as `OK`. It is the same stance the gaussian index takes
against normality tests. With the floor in place the field is still caught
before the smooth is added and the diagnosis is clean after it; with
`min_effect = 0` both are flagged and the verdict stops distinguishing them.

`type = "semivariance"` reports the same information in the form
`nlme::Variogram()` uses, since for standardised residuals the semivariance at
separation `d` is `1 - r(d)`.

### A cycle is not autocorrelation

Both fill the autocorrelation plot with flagged lags. Only one of them is fixed
by an AR term.

The partial autocorrelation is what separates them, and so is the *shape* of the
ordinary one. Autoregression decays through zero and stays there. A cycle comes
back up to a positive peak at its period -- a spike at lag 12 in monthly data,
lag 7 in daily data. `ilm_check_ar()` reads that difference and names the period
rather than reaching for AR(1), which cannot represent a cycle at all.

The remedy is ordinary fixed effects; there is no new likelihood involved, only
the right columns:

```{r seasonterms}
head(round(ilm_fourier(1:12, period = 12, K = 2), 3), 3)
```

```{r seasonfit, eval = FALSE}
# a smooth annual cycle in monthly data
ilm_model(y ~ x + ilm_fourier(month, 12, K = 2) + (1 | subj), data = dd,
          family = "gaussian")

# local flexibility instead, for a long cycle such as day-of-year
ilm_model(y ~ x + ilm_cyclic(doy, 365.25, df = 8) + (1 | subj), data = dd,
          family = "gaussian")
```

`ilm_cyclic()` is a cubic spline that closes on itself: the value and its first
two derivatives match across the wrap, so December runs into January without a
step. An ordinary `bs()` or `ns()` on a phase variable does not know the two
ends of the cycle are the same point, and leaves a jump there.

Which to use is less consequential than it looks. Measured at matched degrees of
freedom on a twelve-phase cycle, Fourier terms and the cyclic spline came within
a few AIC of each other across four shapes. Where the shape genuinely *jumped*,
a plain `factor(month)` beat both by more than a hundred, because no smooth
basis buys a step cheaply. The spline pulls ahead only on long cycles, and only
above about six degrees of freedom, by at most 23 AIC. Pick on the shape you
expect, not on the basis.

## When the spread is not constant

`ilm_check_variance()` reports two patterns, and both have the same remedy:
model the logarithm of the dispersion instead of treating it as one number.

```{r dispformula, eval = FALSE}
# a separate spread per group
ilm_model(y ~ x + (1 | id), data = dd, family = "gaussian",
          dispformula = ~ site)

# spread that grows with the fitted mean; `mu` is the one reserved name
ilm_model(y ~ x + (1 | id), data = dd, family = "gaussian",
          dispformula = ~ mu)
```

These are `nlme`'s `varIdent()` and `varPower()`, and `glmmTMB`'s
`dispformula`, written one way. On a three-group design illume, `glmmTMB` and
`gls` returned the same log-likelihood to five decimals and the same standard
errors; on a power-of-the-mean design illume and `gls` agreed on the power to
two decimals.

It works for any family that has a dispersion to model — the negative
binomial's overdispersion and an accelerated failure time model's scale as well
as a gaussian standard deviation. A Poisson or binomial response has no
dispersion parameter at all, its spread being fixed by its mean, and asking for
one says so.

Two things follow. Exact t and F inference is switched off, because it rests on
a constant variance and the variance is now itself estimated — `gls()` reports
t here by convention, but this package reserves t for where it is exact. And
the fit is warm started from the ordinary one, which matters most for `~ mu`:
from a cold start the fitted mean is zero for every row, `log|mu|` is the log
of the numerical floor, and the power multiplying it can go anywhere. Measured
on data with a true power of 1.0, a cold start converged falsely with the power
at -115818 and the slope collapsed to zero.

## Floors, ceilings and detection limits

A reading of `0` from an assay whose detection limit is 0 does not mean zero.
It means "somewhere at or below the limit", and fitting it as a zero pulls the
mean down and the variance in. Same for a scale that tops out, a score capped
at 100, a measurement beyond an instrument's range.

`ilm_describe()` names the pile-up before you model it:

```{r pileup}
set.seed(1)
y <- 0.5 + rnorm(1500, 0, 1.4)
d0 <- data.frame(unbounded = y, capped = pmin(y, 1.2))
ilm_describe(d0)[, c("variable", "gauss", "gauss_note")]
```

and `ilm_censor()` is the remedy. A censored observation contributes the
probability of the interval it is known to lie in rather than a density at a
value it was never observed at:

```{r tobit, eval = FALSE}
ilm_model(y ~ x + (1 | subj), data = dd, family = "gaussian",
          censor = ilm_censor(dd$y, upper = 1.2))
```

With no random effects this is the Tobit model, and it agrees with
`survival::survreg()` to five decimal places in the coefficients and to 1e-6 in
the log-likelihood. What illume adds is that the random effects still work.
Measured on a 60-unit panel censored at 18%, ignoring the ceiling attenuated the
slope from 1.02 to 0.79 against a truth of 1.0, and the between-unit standard
deviation from 0.78 to 0.65 against a truth of 0.8.

Two things follow automatically. Exact t and F inference is switched off,
because a censored likelihood mixes densities with tail probabilities and is
only asymptotically normal — `survreg()` reports Wald for the same reason. And
the quantile residual of a censored row is drawn across the probability of its
interval, the same randomisation Dunn and Smyth use for a discrete response;
without it every censored row lands on one quantile and the QQ plot shows a
spike at the limit that has nothing to do with fit.

## Time to an event

The same interval-probability mechanism gives parametric survival models. Three
families are available, and all three are the same model with a different error
distribution:

$$\log T = X\beta + \text{scale} \times W$$

`"weibull"`, `"lognormal"` and `"loglogistic"`, for `W` extreme-value, normal
and logistic respectively.

```{r aft, eval = FALSE}
ilm_model(time ~ treatment + age + (1 | centre), data = dd, family = "weibull",
          censor = ilm_surv(dd$time, dd$event))
```

`ilm_surv()` takes `event` in the convention `survival::Surv()` uses — `1` when
the event happened, `0` when the subject was still event-free at the end of
follow-up. It exists because that is the opposite of the code stored
internally, and getting it backwards fits the model to the wrong subjects
without looking wrong.

A coefficient is a **log time ratio**: `+0.5` means survival times are `exp(0.5)`
times longer. That reading holds whatever the baseline hazard does, which is the
appeal of an accelerated failure time model over a proportional hazards one.

All three agree with `survival::survreg()` to about seven decimal places in the
coefficients and ten in the log-likelihood, with the same standard errors. What
illume adds is the random effect. On an 80-centre study with 44% censoring,
a frailty Weibull recovered a slope of 0.746, a scale of 0.597 and a
between-centre standard deviation of 0.465 against truths of 0.7, 0.6 and 0.5;
fitting the same data without the centre term pushed the between-centre spread
into the scale, which came back at 0.729.

### Survival curves, and whether the family was right

```{r survcurve, eval = FALSE}
ilm_plot_survival(fit, dd$time, dd$event)
```

The fitted curve is drawn over the Kaplan-Meier estimate, which assumes nothing
about the shape of the baseline. Where the two part company, the parametric
assumption is what is wrong, and the footnote says to try another family and
compare by AIC.

The grey band is not a confidence interval for the curve. It is the spread of
the Kaplan-Meier across datasets simulated from the fit and refitted, which is
the reference every check in this package uses. It matters here more than most:
on a correctly specified 500-subject model the largest vertical gap between the
two curves was 0.12, which looks alarming and is not — the Kaplan-Meier of 500
subjects wobbles by that much in the tail, and the verdict came back `OK` with
none of the curve outside the band. The same measure on a Weibull fitted to
log-logistic data gave a *smaller* gap, 0.09, and a verdict of `FAIL` with 26%
of the curve outside. A threshold on the gap would have got both backwards.

`ilm_survival()` returns the curve itself, with intervals, for plotting
elsewhere.

Residuals work as they do everywhere else. The quantile residual of a censored
subject is drawn across the probability of the interval they are known to lie
in, so it stays uniform under a correct model — measured across all three
families at 40% censoring, Kolmogorov-Smirnov p-values ran from 0.37 to 0.90.

## Interactions

Interactions are ordinary terms: `x * z` and `g * x` fit, their columns are
grouped into one multi-degree-of-freedom row in `ilm_anova()`, and
`ilm_check_collinearity()` reports them as single terms.

Two things are worth knowing.

**Type III depends on how factors are coded.** With R's default treatment
contrasts, the main-effect row for a variable involved in an interaction tests
its effect *at the reference level* of the factor it interacts with, not
averaged over it. On a 600-row model the `x` row came back at a chi-square of
167 under `contr.treatment` and 1086 under `contr.sum` -- same model, same data.
Only the second is what a Type III main effect is supposed to mean, so
`ilm_anova()` warns and names the fix:

```{r type3, eval = FALSE}
ilm_model(y ~ g * x + (1 | subj), data = dd, family = "gaussian",
          contrasts = list(g = "contr.sum"))
```

`type = 2` sidesteps the question entirely, at the cost of testing a different
(and for main effects, usually more useful) hypothesis.

**An effect plot cannot show one curve.** When a term is in an interaction it has
no single effect, so `ilm_plot_model(fit, "effect")` draws one curve per level of
the interacting factor -- or, for a numeric partner, one per quartile -- and says
so underneath rather than quietly showing you the modal level:

```{r effectinter, eval = FALSE}
ilm_plot_model(fit, "effect", term = "x")
```

## Working with other packages

`illume` fits into the usual ecosystem. `car::Anova()` dispatches to the correct
joint test, `performance::model_performance()` and `check_model()` work, and
`marginaleffects` can be enabled with one call:

```{r ecosystem, eval = FALSE}
ilm_register_marginaleffects()
marginaleffects::avg_slopes(fit, variables = "x1")
```

## Where to read more

The model itself is a baseline-category logit with random effects; Agresti
(2013), *Categorical Data Analysis*, chapter 8, is the standard reference. The
fitting method is described in Kristensen et al. (2016), *Journal of Statistical
Software* 70(5). The mixed-model representation of smooths follows Wood (2017),
*Generalized Additive Models*, section 5.4. The parametric bootstrap follows
Halekoh and Højsgaard (2014), *Journal of Statistical Software* 59(9).
