---
title: "Regression models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2024)
```

`ilm_model()` fits every model in this package. The formula is lme4's, the
response distribution is an argument, and the object that comes back is the
same whatever you fitted -- so `summary()`, `ilm_anova()`, `predict()` and the
diagnostics do not change as the model gets more complicated.

This vignette works through it: a first fit and how to read it, then the
families one at a time, then the structures that go on top of them -- random
slopes, smooths, correlation over time and space, non-constant spread,
censoring. The *Introduction* vignette is the shorter orientation; the
*Workflow* vignette is the map of where this stage sits.

```{r}
library(illume)
```

## Why a multinomial mixed model is fitted this way

Fitting a multinomial mixed model has meant choosing which problem to accept.
Penalised quasi-likelihood (`mclogit`) is fast, and in sparse data it shrinks
fixed effects toward zero -- 27% attenuation with 100 clusters of 4 -- while
reporting standard errors that do not know it, so coverage falls to 0.891
against a nominal 0.95. MCMC (`brms`) is exact up to Monte Carlo error and takes
about 46 seconds of sampling per fit, which is a long time when you are still
deciding what the model is.

illume takes a third route: the Laplace approximation via RTMB, with the random
effects carried as a matrix-normal block. Against brms on the same data,
estimates agree to a mean of 0.10 standard errors with an SE ratio of 0.984;
against mclogit it holds coverage at 0.949 where PQL reaches 0.891. Median fit
time is 0.60 seconds, roughly 75 times faster than brms' sampling alone.

One caveat travels with that: mclogit's RMSE is *lower* (0.336 against 0.409),
because shrinkage buys a variance reduction that more than pays for the bias.
For prediction that is a defensible trade. For a coefficient you intend to
interpret it is not, and coverage is the measure that says so.

On data that misbehave the picture is less tidy. With an outcome category
under 4% of the data, illume *inflates* coefficients rather than shrinking
them, and its mean absolute bias there is worse than mclogit's; and in the
study, with a random-effect SD near zero, only 63.3% of its fits counted as
converged against mclogit's 100%. (That study required a positive definite
Hessian. illume has since reported the fixed effects when a variance sits at
its boundary, and on the same datasets 395 of 400 fits now give them.) The whole
account, including where illume comes off worse, is in
`vignette("benchmarking")`.

## A first model

We will simulate data where an outcome with three categories depends on a
continuous predictor `x1` and a grouping factor `grp`, measured repeatedly within
subjects.

```{r simulate}
n_subj <- 40; per <- 20; N <- n_subj * per
dd <- data.frame(
  subj = factor(rep(seq_len(n_subj), each = per)),
  x1   = rnorm(N),
  grp  = factor(sample(c("a", "b", "c"), N, TRUE))
)

# true structure: an effect of x1 and grp, plus subject-level variation
X  <- model.matrix(~ x1 + grp, dd)
bt <- matrix(c(0.1, -0.2,  0.9, 0.5,  -0.6, 0.4,  0.2, -0.5), 4, 2, byrow = TRUE)
b  <- matrix(rnorm(n_subj * 2), n_subj, 2) %*% diag(c(0.8, 0.6))
eta <- X %*% bt + b[as.integer(dd$subj), ]
P <- exp(eta %*% t(contr.sum(3))); P <- P / rowSums(P)
dd$y <- factor(c("low", "mid", "high")[apply(P, 1, function(p) sample.int(3, 1, prob = p))],
               levels = c("low", "mid", "high"))

table(dd$y)
```

The model is written the way you would write it in `lme4`:

```{r fit}
fit <- ilm_model(y ~ x1 + grp + (1 | subj), data = dd,
                 family = "multinomial", verbose = FALSE)
fit
```

## Reading the output

```{r summary}
summary(fit)
```

Two things in that output deserve attention.

**The coefficients are sum-to-zero contrasts.** A coefficient such as `low:x1` is
the effect of `x1` on the `low` category *relative to the average across all
categories*, not relative to a baseline category. This differs from
`nnet::multinom()` and `brms`, which compare each category against a reference.
The summary prints a reminder, because it is easy to misread if you are used to
the other convention.

**The assumption checks are printed with the coefficients.** That is unusual for
a `summary()` method and it is deliberate, for a reason worth explaining.

## Why the diagnostics are so prominent

This class of model fails *quietly*. A fit can converge, produce entirely
reasonable-looking coefficients with small standard errors and significance
stars, and yet the covariance matrix behind those standard errors can be
unusable -- meaning the uncertainty is meaningless even though nothing in the
output looks wrong.

Every fit therefore carries a table of checks:

```{r checks}
head(fit$checks[, c("check", "status", "detail")], 8)
```

Each check has one of five statuses. `OK` and `FAIL` are self-explanatory. `WARN`
means something is worth a look. `BOUNDARY` means a variance sits at zero, where
the usual tests do not apply; the fit says so as it happens, and names
`boundary = "avoid"`, a small penalty that keeps the variance off zero at a
cost in bias that `?ilm_model` spells out. And `INCONCLUSIVE` means there is not enough
information to judge -- which is reported honestly rather than being passed off
as a clean bill of health.

Anything that is not `OK` also carries a reason and a suggested remedy:

```{r check-detail}
bad <- fit$checks[fit$checks$status != "OK", ]
if (nrow(bad)) bad[1, c("check", "cause", "suggestion")] else "all checks passed"
```

A suggestion in words still has to be turned into code, and `ilm_remedies()`
does that: it lists a remedy for every check that is not `OK`, each written as
the change to `ilm_model()` that makes it, and `ilm_apply_remedy()` refits with
the one you choose. *When there are many categories*, below, shows both.

### The most useful check

The one to look at first is `latent_budget`. A mixed model introduces one
unobserved value per random effect per group. The method used to fit the model
(the Laplace approximation) is accurate when each of those values is informed by
a reasonable amount of data, and degrades when it is not.

Importantly this is a *global* constraint. A single term can look perfectly
healthy on its own and the model can still be over-stretched, because what
matters is the total. Per-term checks cannot see that; this one can.

## Testing fixed effects

`ilm_anova()` gives an analysis of deviance. Each term is tested **jointly
across all category dimensions**, so a predictor with one column is tested on two
degrees of freedom when there are three categories.

```{r anova}
ilm_anova(fit, type = 3)
```

Type III tests each term with all others present; Type II tests each term after
the terms that do not contain it. They agree when there are no interactions. Wald
tests are the default because they need no refitting; a likelihood-ratio version
is available with `test = "LRT"` and is more reliable for very strong effects.

For small samples neither is ideal, because both rely on large-sample
approximations. What to do about it depends on the family.

For a **gaussian** mixed model, `ilm_denom_df()` supplies a finite denominator
degrees of freedom -- Satterthwaite by default, Kenward-Roger where it applies
-- which turns the chi-square into an F:

```{r ddf, eval = FALSE}
ilm_denom_df(fit_gaussian, L)                       # for one contrast
```

Satterthwaite works for every structure this package fits. Kenward-Roger
also inflates the standard errors for the uncertainty in the variance
components, and agrees with `pbkrtest`. It needs a REML fit
(`reml = TRUE`) whose random terms are intercepts and slopes with one
residual variance, and says so rather than returning a number it cannot
stand behind.

For **any other family**, including the multinomial here, neither approximation
is derived and neither is offered. `ilm_pb_lrt()` replaces the assumed
reference distribution with a simulated one instead, which needs no such
derivation:

```{r pb, eval = FALSE}
ilm_pb_lrt(fit, "x1", B = 200, ncores = 4)
```

## Maximum likelihood, and when to switch to REML

Variance components estimated by maximum likelihood are biased downward,
because ML does not account for the degrees of freedom spent on the fixed
effects. With many clusters that hardly matters; with few it carries through
into standard errors and degrees of freedom.

`reml = TRUE` corrects it:

```{r reml, eval = FALSE}
ilm_model(y ~ x1 + grp + (1 | id), data = d, family = "gaussian",
          reml = TRUE)
```

The default is maximum likelihood, and the reason is the order you work in. A
restricted likelihood is the likelihood of contrasts orthogonal to the design
matrix, so changing the fixed effects changes which data it is the likelihood
of. Two REML fits with different fixed effects are not comparable, and their
difference is not a likelihood ratio.

So: **settle the mean structure under the default, then refit with
`reml = TRUE` for the estimates you report.** Once set, `ilm_anova(test =
"LRT")` and `ilm_pb_lrt()` refuse rather than quietly comparing things that
are not comparable, and so does `ilm_robust()` -- a sandwich needs
per-observation scores, and integrating the coefficients out leaves none.

For a gaussian response this is REML exactly: the Laplace approximation to
the restricted likelihood is exact for a linear model, and the variance
components agree with `lme4` to 1e-05. `reml = TRUE` works for every other
family too, as it does in `glmmTMB`, but there the same integral is only
approximately a restricted likelihood. It reduces the downward bias without
REML's exact properties, and `fit$reml_exact` says which of the two a fit has.

`ilm_dag_model()` (for a gaussian response) and `ilm_aov_ez()` default to REML,
because in both the fixed effects were fixed before any data were seen: by the
graph in one case and by the design in the other.

## Predictions

By default `predict()` returns a probability for every observation and category,
as `nnet::multinom(type = "probs")` does.

```{r predict}
head(round(predict(fit), 3))
```

There is a choice to make here, and `groups` makes it. With
`groups = "typical"` (the default) the random effects are set to zero, giving
probabilities for a *typical* subject. With `groups = "population"` the
prediction is averaged over the distribution of subjects, giving probabilities
for the *population*. The subjects' own fitted effects are in `ilm_fitted()`,
for the rows the model was fitted to.

```{r marginal}
grid <- data.frame(x1 = 0, grp = factor(c("a", "b", "c"), levels = levels(dd$grp)))
round(predict(fit, newdata = grid, groups = "typical"), 3)
round(predict(fit, newdata = grid, groups = "population", ndraw = 200), 3)
```

These differ because averaging and the softmax transform do not commute: the
average of the transformed values is not the transform of the average. The
population-averaged probabilities are pulled toward being more even across
categories. Which you want depends on the question you are asking.

Standard errors and intervals come from simulation, and the intervals are
percentile intervals, so they always lie between 0 and 1.

```{r se}
pr <- predict(fit, newdata = grid, se.fit = TRUE, nsim = 100)
round(data.frame(fit = pr$fit[, 1], lower = pr$lower[, 1], upper = pr$upper[, 1]), 3)
```

## Random slopes, smooths and correlated time

Beyond random intercepts, the formula accepts random slopes and penalised
smooths:

```{r more, eval = FALSE}
ilm_model(y ~ x1 + time + (1 + time | subj), data = dd, family = "multinomial")
ilm_model(y ~ x1 + s(xs, k = 10) + (1 | subj), data = dd, family = "multinomial")
ilm_model(y ~ x1 + t2(lon, lat) + (1 | site), data = dd, family = "multinomial")
```

Two rules for smooths: use `t2()` rather than `te()` for tensor products, and
write them unqualified -- `s(x)`, not `mgcv::s(x)` -- because mgcv identifies
smooths by name.

A smooth is named as mgcv names it. `s(x)` is `"s(x)"`, and a smooth with a
numeric `by`, `s(x, by = z)`, is `"s(x):z"`. Its unpenalised columns among the
fixed effects carry the name with `.f1`, `.f2`, ... after it (`"s(x):z.f1"`),
and its penalised part is the random term of that name. A smooth with a
numeric `by` used to be called `"s(x)"`, with columns `"s(x).f1"` and
`"s(x).f2"`. Beside a plain `s(x)` that collided, and the fit lost one of the
two. The section "How smooth terms are named" in `?ilm_model` has the whole
scheme.

A numeric `by` variable is not also entered on its own. `s(x, by = z)` is not
centred, so it already carries `z`'s main effect, and `y ~ z + s(x, by = z)`
stops and says so.

A random term's covariance is set through `re_struct`: a list with one element
per term, named by its grouping variable, each element itself a list. Two
settings go in it. `d_cor = FALSE` drops the correlation between a random
intercept and its slopes, in any family. `type` sets the covariance across a
multinomial outcome's categories -- `"us"` by default, `"diag"`, or `"rr"` with
a `rank` -- and matters only there, since every other family has a single
linear predictor:

```{r restruct, eval = FALSE}
## a slope for time, uncorrelated with the subject's intercept
ilm_model(y ~ x1 + time + (1 + time | subj), data = dd, family = "gaussian",
          re_struct = list(subj = list(d_cor = FALSE)))
## a multinomial outcome whose site effects share one dimension
ilm_model(y ~ x1 + (1 | site), data = dd, family = "multinomial",
          re_struct = list(site = list(type = "rr", rank = 1)))
```

A term `re_struct` does not name keeps the default. "When there are many
categories", below, says when to reach for a reduced rank.

## Simple models, and exact inference

Nothing requires a random effect. `ilm_model()` fits an ordinary regression, and
when the family is gaussian and nothing is integrated out it reports **exact**
inference rather than large-sample approximations:

```{r lmlike}
set.seed(1)
sd2 <- data.frame(x = rnorm(200), z = factor(sample(c("a", "b", "c"), 200, TRUE)))
sd2$v <- 2 + 1.5 * sd2$x - 0.8 * (sd2$z == "b") + rnorm(200, 0, 1.2)

f_lm <- ilm_model(v ~ x + z, data = sd2, family = "gaussian", verbose = FALSE)
round(ilm_coef_table(f_lm), 5)
round(summary(lm(v ~ x + z, data = sd2))$coefficients, 5)
```

The two agree to numerical precision, including the residual standard deviation
and the degrees of freedom. `ilm_anova()` likewise switches from chi-square to F:

```{r lmanova}
ilm_anova(f_lm, type = 3)
```

This matters because the chi-square test treats the residual variance as known,
which it is not, making it anti-conservative in small samples. Where an exact
reference exists, illume uses it.

It applies only where it is genuinely available. A Poisson or binomial model
without random effects is still a GLM, with no exact small-sample analogue, so
it keeps z and chi-square. So does any model with a random effect.

## Other families

The same formula syntax, methods and diagnostics apply to every family. What
changes is the response, the link, and whether there is a dispersion parameter:

```{r families, eval = FALSE}
ilm_model(count ~ x + (1 | site), data = dd, family = "poisson")
ilm_model(count ~ x + (1 | site), data = dd, family = "nbinom")   # overdispersed
ilm_model(hit   ~ x + (1 | site), data = dd, family = "binomial")
ilm_model(score ~ x + (1 | site), data = dd, family = "gaussian")
```

### Leaving the family to the response

`family` defaults to `"auto"`, which reads the family off the response and
says which it chose and on what evidence:

```{r family-auto, eval = FALSE}
ilm_model(visits ~ x + (1 | site), data = dd)
#> ilm_model(): family = "poisson", inferred from `visits`: whole numbers from
#> 0 to 14, a count. If the counts are more spread out than a Poisson allows
#> (ilm_check_dispersion() tests it), use family = "nbinom"; if they are
#> ratings rather than counts, an ordered factor with family = "ordinal".
#> Pass `family` to choose another.
```

A factor with two levels, a logical or a 0/1 variable is binomial; an
unordered factor with three or more levels is multinomial and an ordered one
ordinal; values strictly between 0 and 1 are beta. Whole numbers are the hard
case, because a count and a measurement recorded to the nearest unit look
alike: whole numbers that reach down to 0 or 1 are read as a count, and ones
that never come near zero -- a blood pressure, a height -- as a measurement.
The range goes into the message either way, so a misreading shows at a glance.

What the response cannot settle is asked about rather than guessed: a
proportion that touches 0 or 1, or comes with weights that may be numbers of
trials; a two-valued variable coded other than 0/1; a survival time, where the
family is a modelling decision; a date. The choice is written into
`fit$call`, so anything that refits the model keeps it instead of guessing
again on simulated data, and `summary()` marks the family as inferred. It is
a starting point, not a verdict -- a count that is overdispersed is still a
count, and `family = "nbinom"` is a choice only the diagnostics can prompt.

`summary()` reports a dispersion parameter where the family has one -- the
residual standard deviation for gaussian, the overdispersion parameter for
negative binomial -- and omits that section otherwise. Coefficients carry no
category prefix outside the multinomial, and the sum-to-zero note disappears,
because there is only one linear predictor.

A two-category outcome can be stored however your data happen to hold it --
`0`/`1`, a two-level factor, `TRUE`/`FALSE`, or a character column read straight
from a csv. All four give the same fit. For a factor or character response the
*second* level is the one being modelled, exactly as in `glm()`, so
`factor(c("no", "yes"))` models the probability of `"yes"`. If you want the other
direction, relevel the factor. You do not have to remember which way round it is:
`summary()` states the category it is modelling, and the reference it is measured
against.

Two categories belong to `"binomial"` and three or more to `"multinomial"`;
each family says so if it is given the other one's data.

If your counts vary more than Poisson allows, `"nbinom"` adds one parameter `k`,
giving variance `mu + mu^2/k`. Smaller `k` means more overdispersion; large `k`
approaches the Poisson.

Excess zeros have a part of their own: `ziformula = ~ 1` adds a zero-inflation
probability, and `zi_type = "hurdle"` separates whether the response clears
zero from how far past it goes -- see `?ilm_model` for which is which, and
`ilm_check_zeros()` for whether either is needed. Proportions strictly
between 0 and 1 take `family = "beta"`, ordered categories the ordinal
families, and times to an event the survival families below. Tweedie is out
of scope; `glmmTMB` covers it well.

## When there are many categories

The number of covariance parameters grows quickly with the number of categories.
With an unstructured covariance and `J` categories you need `(J-1)J/2` parameters
*per grouping factor* -- 3 when `J = 3`, but 45 when `J = 10`. Many designs
cannot support that.

The remedy is a reduced-rank structure, which describes the covariance with fewer
dimensions:

```{r rr, eval = FALSE}
ilm_model(y ~ x1 + (1 | subj), data = dd, family = "multinomial",
          re_struct = list(subj = list(type = "rr", rank = 3)))
```

This saves parameters *and* latent values, and the latter is usually the bigger
gain. When a structure is too rich for the data, the checks say so and name a
specific rank to try. Five categories and twenty subjects is such a case: an
unstructured covariance for `subj` has 10 parameters, and there are two
subjects for each of them.

```{r many-cats}
set.seed(7)
d5 <- data.frame(subj = factor(rep(1:20, each = 15)), x1 = rnorm(300))
# the subjects differ along one dimension, shared by the four linear predictors
eta5 <- outer(0.4 * d5$x1, seq(1, -1, length.out = 4)) +
  outer(rnorm(20)[d5$subj], seq(0.8, -0.8, length.out = 4))
P5 <- exp(eta5 %*% t(contr.sum(5))); P5 <- P5 / rowSums(P5)
d5$y <- factor(apply(P5, 1, function(p) sample(letters[1:5], 1, prob = p)))

f5 <- ilm_model(y ~ x1 + (1 | subj), data = d5, family = "multinomial",
                verbose = FALSE)
rem <- ilm_remedies(f5)
rem
```

Each remedy is the change that makes it, and its tier says what applying it
would change: `structural` alters the covariance and leaves the fixed effects
meaning what they did, while `estimand` would change what they estimate.
Three remedies name a rank, because three checks ask three questions:
`re_levels` what twenty subjects can support, `latent_budget` what the whole
model can, and `sigma_rank` how many dimensions of the estimated covariance
have any variance. The first is fixed by the design before anything is
estimated, so it is the one to try, and the refit's own checks say whether it
was enough:

```{r many-cats-apply}
f5r <- ilm_apply_remedy(f5, rem, rem$id[startsWith(rem$check, "re_levels")][1])
f5r$checks[f5r$checks$status != "OK", c("check", "status", "detail")]
```

The level check now warns rather than fails -- four parameters from twenty
subjects is five subjects each, short of the six it asks for -- and the
covariance no longer sits at its boundary. The refit went through the model's
own call, so everything else about it is as it was, and `f5r$remedy_log`
records how it was reached.

## Checking the fitted model

`ilm_appraise()` draws diagnostic panels built for a categorical outcome rather
than borrowed from tools designed for continuous responses:

```{r appraise, fig.width = 7, fig.height = 5, eval = FALSE}
ilm_appraise(fit)
```

Two further checks are worth knowing about, because broad residual summaries have
very little power to detect the problems they address. `ilm_check_omitted()`
tests variables you did *not* include, and `ilm_check_ar()` tests for leftover
correlation over time:

```{r targeted, eval = FALSE}
ilm_check_omitted(fit, dd)
ilm_check_ar(fit, time = dd$time, group = dd$subj)
```

And `ilm_consistency()` asks the most direct question available: can the model
recover itself from data it generated? If not, the fitting method is not reliable
for this model and this amount of data.

```{r consistency, eval = FALSE}
ilm_consistency(fit, B = 50, ncores = 4)
```

## Structure over time

Repeated measurements on the same unit are rarely independent, and there are two
quite different reasons why -- which matter because they need opposite remedies.

`ilm_plot_acf()` draws the residual autocorrelation and partial autocorrelation
by lag, within group, with a verdict on each lag:

```{r acfplot, eval = FALSE}
ilm_plot_acf(fit, time = dd$time, group = dd$subj, maxlag = 14)
```

Three things make this not `stats::acf()`:

- Pairs are matched on exact time differences **within group**. A panel dataset
  stacked long is many short series, and pairing across the joins manufactures
  correlation that is not there.
- The band is not `2/sqrt(n)`. Residuals within a group sum to roughly zero, so
  pairs of them correlate *negatively* even when the model is exactly right --
  measured, around -0.07 at every lag on a 60-unit panel. A band centred on zero
  calls correct models wrong. The band here comes from simulating from the fit
  and refitting.
- The band is drawn at the critical value the p-value is computed against, so a
  point outside the band is exactly a lag the check flags. The picture and the
  table cannot disagree.

### When the times are not evenly spaced

`ilm_ar1()` needs a common step, because "one step" has to mean something.
When visits fall where they fall, use `ilm_car1()`: the correlation between two
observations is `rho` raised to the gap between them, which is `nlme`'s
`corCAR1()` and reduces to AR(1) when every gap is one.

```{r car1, eval = FALSE}
ilm_model(y ~ x + (1 | subj), data = dd, family = "gaussian",
          ar = ilm_car1(~ day | subj))
```

Written as `~ day | subj`, the time and the unit are read from the data after
incomplete rows are dropped, so they cannot fall out of step with the
response, as vectors taken from the whole data frame can. All three
structures over time take this form.

Both constructors report their **latent budget** before you fit, because that is
the constraint that bites. These are latent-variable structures: one latent
value exists per distinct unit-and-time, so genuinely continuous times give one
latent per observation. For a gaussian response that is fine — the model is
linear-Gaussian, the Laplace approximation is exact, and coverage measured over
400 replicates at exactly one observation per latent came back at 0.955, 0.937
and 0.949 against a nominal 0.95. For a categorical response it is not, and the
fit says so. Rounding `time` to a coarser grid, so observations share a latent,
is the usual fix.

### A level that drifts

A CAR(1) process wanders and comes back: however far it strays, it is pulled
towards its mean. A level that drifts and stays where it drifted to is a
random walk, which `ilm_rw1()` fits -- the local-level model of time-series
analysis. Each unit's walk is zero at its first time, so the fixed effects
give its starting level, and from there it moves by changes whose variance
is proportional to the time elapsed:

```{r rw1, eval = FALSE}
ilm_model(y ~ x, data = dd, family = "gaussian",
          ar = ilm_rw1(~ day | subj))
```

Add `(1 | subj)` when units start from different levels. For a gaussian
response the fit is the exact likelihood of the local-level model, by
maximum likelihood or by REML. `ilm_cells()` lists the fitted walk's cells:
each unit's times, which is its last, and where the fit holds each value.

### Checking a correlation structure over irregular time

`ilm_check_ar()` matches pairs at exact lags, which stops working once the times
are irregular: on a sixty-unit panel with six observations each drawn from
thirty possible times, it found 53 pairs at lag 1, five at lag 2 and none
beyond. `ilm_variogram()` bins every within-group pair by how far apart it is
instead, and uses all 900.

```{r variogram, eval = FALSE}
ilm_variogram(fit, time = dd$day, group = dd$subj)
```

It distinguishes three shapes, because they need three different remedies:

| what you see | what it is | what to do |
|---|---|---|
| decays from the shortest separation | autoregression | `ilm_car1()` |
| the same level at every separation | a group effect | add or widen a random effect |
| alternates in sign | a cycle | `ilm_fourier()`, `ilm_cyclic()` |

The envelope again comes from refitting simulated data, and again is not centred
on zero: residuals within a group sum to roughly zero, so pairs of them
correlate negatively even under a correct model, and more so the further apart
they are.

### Correlation in space

The same function takes coordinates instead of a time, because separation in
space and separation in time are the same statistic over a different distance:

```{r spatialvg, eval = FALSE}
ilm_variogram(fit, coords = dd[, c("easting", "northing")])
```

The remedy it names is different, and deliberately so. illume fits no spatial
covariance — no Matern field, no exponential one — so the advice does not
pretend otherwise. What it points at is putting the structure in the **mean**:
a tensor-product smooth of the coordinates, `t2(x, y)`, absorbs smooth spatial
variation, and a random effect for a spatial grouping absorbs the coarse kind.

Measured on 300 points carrying an exponential field, residual correlation ran
0.242 at the shortest distance down through 0.071 to −0.072 at the longest.
Adding `t2(sx, sy)` cut the first of those to −0.033, dropped AIC from 1165.7 to
1069.4, and narrowed the standard error on the covariate of interest from 0.100
to 0.083.

### A departure too small to act on

Each bin rests on thousands of pairs, so the envelope narrows until any
imperfection clears it. In the example above the residual correlation after the
smooth was about 0.03 against an envelope 0.02 wide — a seventh of what it had
been, and nothing anyone would act on — and four of five bins were still
flagged.

`min_effect` is the floor, 0.1 by default: a bin outside the envelope by less
than that is reported as `OK`. It is the same stance the gaussian index takes
against normality tests. With the floor in place the field is still caught
before the smooth is added and the diagnosis is clean after it; with
`min_effect = 0` both are flagged and the verdict stops distinguishing them.

`type = "semivariance"` reports the same information in the form
`nlme::Variogram()` uses, since for standardised residuals the semivariance at
separation `d` is `1 - r(d)`.

### A cycle is not autocorrelation

Both fill the autocorrelation plot with flagged lags. Only one of them is fixed
by an AR term.

The partial autocorrelation is what separates them, and so is the *shape* of the
ordinary one. Autoregression decays through zero and stays there. A cycle comes
back up to a positive peak at its period -- a spike at lag 12 in monthly data,
lag 7 in daily data. `ilm_check_ar()` reads that difference and names the period
rather than reaching for AR(1), which cannot represent a cycle at all.

The remedy is ordinary fixed effects; there is no new likelihood involved, only
the right columns:

```{r seasonterms}
head(round(ilm_fourier(1:12, period = 12, K = 2), 3), 3)
```

```{r seasonfit, eval = FALSE}
# a smooth annual cycle in monthly data
ilm_model(y ~ x + ilm_fourier(month, 12, K = 2) + (1 | subj), data = dd,
          family = "gaussian")

# local flexibility instead, for a long cycle such as day-of-year
ilm_model(y ~ x + ilm_cyclic(doy, 365.25, df = 8) + (1 | subj), data = dd,
          family = "gaussian")
```

`ilm_cyclic()` is a cubic spline that closes on itself: the value and its first
two derivatives match across the wrap, so December runs into January without a
step. An ordinary `bs()` or `ns()` on a phase variable does not know the two
ends of the cycle are the same point, and leaves a jump there.

Which to use is less consequential than it looks. Measured at matched degrees of
freedom on a twelve-phase cycle, Fourier terms and the cyclic spline came within
a few AIC of each other across four shapes. Where the shape genuinely *jumped*,
a plain `factor(month)` beat both by more than a hundred, because no smooth
basis buys a step cheaply. The spline pulls ahead only on long cycles, and only
above about six degrees of freedom, by at most 23 AIC. Pick on the shape you
expect, not on the basis.

## When the spread is not constant

`ilm_check_variance()` reports two patterns, and both have the same remedy:
model the logarithm of the dispersion instead of treating it as one number.

```{r dispformula, eval = FALSE}
# a separate spread per group
ilm_model(y ~ x + (1 | id), data = dd, family = "gaussian",
          dispformula = ~ site)

# spread that grows with the fitted mean; `mu` is the one reserved name
ilm_model(y ~ x + (1 | id), data = dd, family = "gaussian",
          dispformula = ~ mu)
```

These are `nlme`'s `varIdent()` and `varPower()`, and `glmmTMB`'s
`dispformula`, written one way. On a three-group design illume, `glmmTMB` and
`gls` returned the same log-likelihood to five decimals and the same standard
errors; on a power-of-the-mean design illume and `gls` agreed on the power to
two decimals.

It works for any family that has a dispersion to model — the negative
binomial's overdispersion and an accelerated failure time model's scale as well
as a gaussian standard deviation. A Poisson or binomial response has no
dispersion parameter at all, its spread being fixed by its mean, and asking for
one says so.

Two things follow. Exact t and F inference is switched off, because it rests on
a constant variance and the variance is now itself estimated — `gls()` reports
t here by convention, but this package reserves t for where it is exact. And
the fit is warm started from the ordinary one, which matters most for `~ mu`:
from a cold start the fitted mean is zero for every row, `log|mu|` is the log
of the numerical floor, and the power multiplying it can go anywhere. Measured
on data with a true power of 1.0, a cold start converged falsely with the power
at -115818 and the slope collapsed to zero.

## Floors, ceilings and detection limits

A reading of `0` from an assay whose detection limit is 0 does not mean zero.
It means "somewhere at or below the limit", and fitting it as a zero pulls the
mean down and the variance in. Same for a scale that tops out, a score capped
at 100, a measurement beyond an instrument's range.

`ilm_describe()` names the pile-up before you model it:

```{r pileup}
set.seed(1)
y <- 0.5 + rnorm(1500, 0, 1.4)
d0 <- data.frame(unbounded = y, capped = pmin(y, 1.2))
ilm_describe(d0)[, c("variable", "gauss", "gauss_note")]
```

and `ilm_censor()` is the remedy. A censored observation contributes the
probability of the interval it is known to lie in rather than a density at a
value it was never observed at:

```{r tobit, eval = FALSE}
ilm_model(y ~ x + (1 | subj), data = dd, family = "gaussian",
          censor = ilm_censor(dd$y, upper = 1.2))
```

With no random effects this is the Tobit model, and it agrees with
`survival::survreg()` to five decimal places in the coefficients and to 1e-6 in
the log-likelihood. What illume adds is that the random effects still work.
Measured on a 60-unit panel censored at 18%, ignoring the ceiling attenuated the
slope from 1.02 to 0.79 against a truth of 1.0, and the between-unit standard
deviation from 0.78 to 0.65 against a truth of 0.8.

Two things follow automatically. Exact t and F inference is switched off,
because a censored likelihood mixes densities with tail probabilities and is
only asymptotically normal — `survreg()` reports Wald for the same reason. And
the quantile residual of a censored row is drawn across the probability of its
interval, the same randomisation Dunn and Smyth use for a discrete response;
without it every censored row lands on one quantile and the QQ plot shows a
spike at the limit that has nothing to do with fit.

## Time to an event

The same interval-probability mechanism gives parametric survival models. Three
families are available, and all three are the same model with a different error
distribution:

$$\log T = X\beta + \text{scale} \times W$$

`"weibull"`, `"lognormal"` and `"loglogistic"`, for `W` extreme-value, normal
and logistic respectively.

```{r aft, eval = FALSE}
ilm_model(time ~ treatment + age + (1 | centre), data = dd, family = "weibull",
          censor = ilm_surv(dd$time, dd$event))
```

`ilm_surv()` takes `event` in the convention `survival::Surv()` uses — `1` when
the event happened, `0` when the subject was still event-free at the end of
follow-up. It exists because that is the opposite of the code stored
internally, and getting it backwards fits the model to the wrong subjects
without looking wrong.

A coefficient is a **log time ratio**: `+0.5` means survival times are `exp(0.5)`
times longer. That reading holds whatever the baseline hazard does, which is the
appeal of an accelerated failure time model over a proportional hazards one.

All three agree with `survival::survreg()` to about seven decimal places in the
coefficients and ten in the log-likelihood, with the same standard errors. What
illume adds is the random effect. On an 80-centre study with 44% censoring,
a frailty Weibull recovered a slope of 0.746, a scale of 0.597 and a
between-centre standard deviation of 0.465 against truths of 0.7, 0.6 and 0.5;
fitting the same data without the centre term pushed the between-centre spread
into the scale, which came back at 0.729.

### Survival curves, and whether the family was right

```{r survcurve, eval = FALSE}
ilm_plot_survival(fit, dd$time, dd$event)
```

The fitted curve is drawn over the Kaplan-Meier estimate, which assumes nothing
about the shape of the baseline. Where the two part company, the parametric
assumption is what is wrong, and the footnote says to try another family and
compare by AIC.

The grey band is not a confidence interval for the curve. It is the spread of
the Kaplan-Meier across datasets simulated from the fit and refitted, which is
the reference every check in this package uses. It matters here more than most:
on a correctly specified 500-subject model the largest vertical gap between the
two curves was 0.12, which looks alarming and is not — the Kaplan-Meier of 500
subjects wobbles by that much in the tail, and the verdict came back `OK` with
none of the curve outside the band. The same measure on a Weibull fitted to
log-logistic data gave a *smaller* gap, 0.09, and a verdict of `FAIL` with 26%
of the curve outside. A threshold on the gap would have got both backwards.

`ilm_survival()` returns the curve itself, with intervals, for plotting
elsewhere.

Residuals work as they do everywhere else. The quantile residual of a censored
subject is drawn across the probability of the interval they are known to lie
in, so it stays uniform under a correct model — measured across all three
families at 40% censoring, Kolmogorov-Smirnov p-values ran from 0.37 to 0.90.

## Interactions

Interactions are ordinary terms: `x * z` and `g * x` fit, their columns are
grouped into one multi-degree-of-freedom row in `ilm_anova()`, and
`ilm_check_collinearity()` reports them as single terms.

Two things are worth knowing.

**Type III depends on how factors are coded.** With R's default treatment
contrasts, the main-effect row for a variable involved in an interaction tests
its effect *at the reference level* of the factor it interacts with, not
averaged over it. On a 600-row model the `x` row came back at a chi-square of
167 under `contr.treatment` and 1086 under `contr.sum` -- same model, same data.
Only the second is what a Type III main effect is supposed to mean, so
`ilm_anova()` warns and names the fix:

```{r type3, eval = FALSE}
ilm_model(y ~ g * x + (1 | subj), data = dd, family = "gaussian",
          contrasts = list(g = "contr.sum"))
```

`type = 2` sidesteps the question entirely, at the cost of testing a different
(and for main effects, usually more useful) hypothesis.

**An effect plot cannot show one curve.** When a term is in an interaction it has
no single effect, so `ilm_plot_model(fit, "effect")` draws one curve per level of
the interacting factor -- or, for a numeric partner, one per quartile -- and says
so underneath rather than quietly showing you the modal level:

```{r effectinter, eval = FALSE}
ilm_plot_model(fit, "effect", term = "x")
```

## From association to effect

Everything above estimates associations. Three designs turn one into a causal
claim, each naming an assumption and letting it be partly checked:

```{r causal, eval = FALSE}
# the regression a causal graph implies, with the graph itself tested
g <- ilm_dag("dag { x [exposure] ; y [outcome] ; z -> x -> y ; z -> y }")
ilm_dag_model(g, d)

# a treatment that switches on for some units at some time
ilm_did(panel, "y", "unit", "time", treated = "treated", post = "post")

# a treatment assigned by crossing a threshold
ilm_rdd(d, "score", "running", cutoff = 0)
```

The rule that makes these safe to automate: the **design** fixes the mean
structure and nothing searches over it. Only the error structure is adjusted
when a diagnostic asks. `vignette("causal-models")` covers all three.

## Reading a fit back in words

`ilm_interpret()` writes out what a model says -- each effect on the scale the
response is measured on, how strong the evidence is, what the diagnostics found,
and how far to trust the estimates:

```{r interpret, eval = FALSE}
ilm_interpret(fit)
```

Each effect is said over a change a reader can picture: across the middle half
of a numeric predictor, level by level for a factor, with the model's predictions
at both ends, averaged over the rows it was fitted to -- "across the middle half
of age, predicted income is 49.4 at 29 against 64.7 at 54: 15.2 higher (95%
interval 12.3 to 18.1)". A coefficient of 0.61 says nothing until the reader
knows what a unit of age is and how much age varies; the middle half answers
both. The strength of evidence is the term's joint test, so a factor with
several levels, or a predictor in a multinomial model, gets one verdict.

The prose is templated rather than generated, so the same fit gives the same
words. It says nothing about bias that a diagnostic did not measure, and it uses
causal language only when the object carries a design that licenses it --
otherwise every effect is "associated with".

`ilm_ame()` gives the piece that makes this readable for a model with a link
function: the average marginal effect on the response scale. An odds ratio of 2
can be a 3-point change in probability or a 20-point one depending on where the
data sit, so the change in probability is what gets reported.

```{r ame, eval = FALSE}
ilm_ame(fit)
```

## Ordered outcomes

An ordered response treated as unordered throws the ordering away and pays for
it in parameters: a multinomial spends `J - 1` coefficients per predictor where
a cumulative link model spends one. Treated as a number it invents a spacing the
categories do not have -- "agree" is not twice "neutral".

```{r ordinal, eval = FALSE}
fit <- ilm_model(rating ~ dose + age, data = dd, family = "ordinal")
ilm_thresholds(fit)
```

The model is `P(Y <= j) = F(threshold_j - eta)`, so the linear predictor is
**subtracted** from each cut and a positive coefficient pushes probability
towards the **higher** categories. Reading it the other way reverses every
conclusion, which is why `summary()` prints the convention beside the
thresholds. `"ordinal_probit"` and `"ordinal_cloglog"` change the link.

One coefficient describing every cut is the proportional-odds assumption, and
it is an assumption:

```{r ordinal-check, eval = FALSE}
ilm_check_proportional(fit)
```

If a term acts differently at different cuts, the remedy is
`family = "multinomial"`, which spends more coefficients and assumes nothing
about the ordering. It nests this model, so `AIC()` says whether the extra
parameters earn their place -- and the ordering is real information, so giving
all of it up to accommodate one predictor may cost more than it buys.

## Counts with too many zeros

More zeros than a count model expects is the one misspecification a residual
plot cannot show, because a zero is a perfectly ordinary value of a count.

```{r zeros, eval = FALSE}
ilm_check_zeros(fit)
fit <- ilm_model(visits ~ age, data = dd, family = "poisson", ziformula = ~ insured)
ilm_zi_coef(fit)
```

`zi_type` picks between two models that are **not** the same model. `"inflated"`
is a mixture: some units are structural zeros and the count part can produce
zeros of its own, so a zero has two possible origins and the probability is the
structural share only. `"hurdle"` is two processes: whether the response clears
zero, and how far past it goes, the second fitted to a count that cannot be
zero. Which applies is a question about the subject -- whether there is a
subpopulation that was never at risk -- rather than one the fit can settle,
though `AIC()` does separate them.

## Proportions

A proportion that is not a count of anything -- percent cover, share of time, a
score already scaled to the unit interval -- has no denominator for a binomial
fit, and a gaussian one puts mass outside `[0, 1]` and assumes a spread that
does not shrink at the ends.

```{r beta, eval = FALSE}
fit <- ilm_model(cover ~ elevation, data = dd, family = "beta")
```

The parameterisation is a mean and a **precision**, so a larger dispersion
parameter means *less* spread -- the opposite of every other one here.
`dispformula` lets the precision depend on a covariate.

Values sitting exactly at 0 or 1 have no likelihood, because the density has no
mass there, and the error says so along with the two honest responses: they are
a separate process, in which case model them as one with `ziformula`, or they
are rounding, in which case `ilm_squeeze()` applies the Smithson-Verkuilen shift
and reports how far it moved things.

## When a regressor is not exogenous

A treatment people chose rather than were assigned, a price set in response to
demand, an exposure measured with error. If an instrument is available --
something that moves the regressor without acting on the outcome any other way
-- `ilm_iv()` uses it.

```{r iv, eval = FALSE}
fit <- ilm_iv(earnings ~ schooling + age | distance + age, data = dd)
```

Everything to the right of the bar is the **full** list of exogenous variables,
instruments and covariates alike. Leaving a covariate out does not mean it is
fine, it means it is endogenous with no instrument, and that is an error rather
than a silent one.

Three diagnostics print unasked: the first-stage F, against both the familiar
threshold of 10 and the roughly 104 a conventional 5% t-test actually needs;
Durbin-Wu-Hausman, asking whether the regressor was endogenous at all, since if
it was not then least squares was unbiased and far more precise; and Sargan's J
when there is more than one instrument. When the instrument is weak,
`ilm_iv_ar()` gives a confidence set that stays valid however weak it is -- and
can come back unbounded, which is the honest answer a Wald interval hides.

## A complex sample

Sampling weights are not replicate counts. Handed to a model that treats weights
as frequencies they give standard errors too small by at least
`sqrt(n / sum(w))`, because the likelihood believes it saw `sum(w)`
observations.

```{r survey, eval = FALSE}
des <- ilm_design(dd, weights = ~ wt, ids = ~ psu, strata = ~ stratum)
fit <- ilm_model(y ~ x, data = dd, family = "binomial", design = des)
ilm_svy_coef(fit)
```

The coefficients are fine in a fixed-effects fit -- a weighted likelihood is
design-consistent for the population parameter -- which is what makes the
mistake easy to miss: everything looks right except the uncertainty, and the
uncertainty looks better than right. The degrees of freedom are the number of
sampling units less the number of strata, not the number of rows.

## Mediation

How much of an effect runs **through** something else.

```{r mediate, eval = FALSE}
fm <- ilm_model(stress ~ treat + age, data = dd, family = "gaussian")
fy <- ilm_model(outcome ~ treat * stress + age, data = dd, family = "gaussian")
md <- ilm_mediate(fm, fy, treat = "treat", mediator = "stress")
ilm_mediate_sens(md)
```

The Baron-Kenny product of coefficients is the special case where the outcome
model is linear and has no treatment-by-mediator interaction. With an
interaction the indirect effect differs by arm and one product cannot be two
numbers; with a binary outcome the product is not on the scale of the effect at
all. `ilm_mediate()` estimates the counterfactual definitions, and reproduces
the product exactly where the product is right.

The decomposition needs no unmeasured confounding of the **mediator** and the
outcome, which the data cannot check because the mediator was not randomised.
`ilm_mediate_sens()` asks the answerable question instead: how strong would such
confounding have to be before the indirect effect went away, and how that
compares with the predictors you did measure.

## Cluster-robust standard errors

When the mean structure is credible but the variance structure is not, and you
would rather not say what the right one is:

```{r robust, eval = FALSE}
ilm_robust(fit, ~ school)
```

The defaults are CR2 with a t reference on Bell-McCaffrey degrees of freedom,
because that is what covers: with a cluster-level predictor and 10 clusters,
the usual CR0-with-a-normal-reference covers a nominal 95% interval 80% of the
time and this covers 95%. Those degrees of freedom are far below the number of
clusters -- a treatment assigned to three schools of thirty does not have
twenty-nine behind it -- and reporting them is the warning.
