---
title: "The illume workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The illume workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This is a map rather than a tutorial. It walks one analysis from a sample-size
calculation through to reporting, eleven stages, with a short paragraph and a
minimal block for each, and a pointer to the vignette or help page that covers
it properly.

Nothing here is compulsory. Most analyses skip several stages, and a few use
only one. The point of the map is that when you do need a stage, the function
for it is in the same package, speaks the same object, and names its own
remedy when it finds a problem.

```{r}
library(illume)
```

---

## 1. Power

Before the data exist. `ilm_power()` treats a model as the truth, simulates
studies at the sizes you are considering, refits each, and counts how often the
effect is found. It works for any family the package fits, because it never
needs a closed-form variance.

```{r}
pilot <- ilm_model(y ~ dose, data = pilot_data, family = "binomial")
pw <- ilm_power(pilot, n = c(200, 400, 800), effect = 0.4, sims = 500)
ilm_power_n(pw, target = 0.8)
plot(pw)
```

Two numbers it reports that most software does not: the **Monte Carlo interval**
around the power estimate, because 0.80 from 200 replicates is really 0.74 to
0.86 and a required sample size read off that curve is a range; and the
**convergence rate**, because a replicate that failed to fit has not detected
anything, and dividing only by the ones that worked gives power conditional on
convergence, which is a different and more flattering quantity.

With no pilot to fit, `ilm_power_design()` takes the assumptions instead --
the design, the expected cell means, the residual SD and the ICC -- and
`ilm_scaffold()` returns the assumed model on its own, so what it implies can
be checked before a sample size is quoted from it.

```{r}
d <- list(arm = c("control", "treatment"), time = c("pre", "post"))
pw <- ilm_power_design(change ~ arm * time + (1 | id), design = d,
                       within = "time", n_unit = c(60, 120, 240, 400),
                       cells = c(control.pre = 20, control.post = 21,
                                 treatment.pre = 20, treatment.post = 25),
                       sd = 6, icc = 0.5, term = "arm:time", sims = 300)
ilm_power_n(pw, target = 0.8)
#>  effect   n  n_lower n_upper n_unit n_unit_lower n_unit_upper
#>       4 330 292.5298 365.872    165     146.2649      182.936
```

`n_unit` is participants and `n` is rows: 165 people measured twice.

See `?ilm_power`, `?ilm_scaffold` and the *Effect size and power* vignette.

---

## 2. Exploration

Stages 2 to 4 use `illumex`, the exploration package `illume` attaches, so
nothing extra needs loading. Descriptive statistics, plots, and bootstrap
intervals for the quantities you are about to model. `ilm_describe_all()`
summarises every column at once;
`ilm_plot_all()` picks a sensible plot per variable; `ilm_boot_ci()` and
`ilm_boot_diff()` put intervals on means, medians and differences between
groups without assuming normality.

```{r}
ilm_describe_all(d)
ilm_plot_all(d)
ilm_boot_diff(score ~ arm, data = d)        # every pair, simultaneous by default
```

See the *Exploring data* vignette.

---

## 3. Structure and profiling

When the columns are many and correlated, `ilm_reduce()` finds the few
directions they share -- PCA, MCA or FAMD depending on the types, or a
generalized low rank model with a loss per column type -- `ilm_cluster()`
groups the rows in that space, and `ilm_profile()` does both and describes what
distinguishes each group.

```{r}
pr <- ilm_profile(d, ndim = 5)               # closed-form FAMD
pr <- ilm_profile(d, ndim = 5, method = "glrm")   # a loss per column type
ilm_plot_profile(pr)
```

See the *Profiling* vignette and `?ilm_glrm`.

---

## 4. Outliers and anomalies

Two different questions. `ilm_outliers()` asks whether a value is extreme in
its own column. `ilm_anomaly()` asks whether a **row** is a plausible
combination: someone 150 cm tall is unremarkable, someone weighing 110 kg is
unremarkable, and someone who is both is not, and nothing in either column's
distribution says so.

```{r}
ilm_outliers_all(d)                          # one column at a time
ilm_anomaly(d)                               # the combination
```

`ilm_anomaly()` also names the column driving each flagged row, because an
anomaly nobody can explain is not actionable.

See the *Anomaly detection* vignette.

---

## 5. Missing data

`ilm_check_missing()` asks the question that matters, which is not how much is
missing but whether it is missing in a way that biases the answer: complete
cases stay unbiased when missingness is unrelated to the **outcome given the
predictors**, which is weaker than missing-completely-at-random and is
testable. When it is not, `ilm_impute()` is the remedy.

```{r}
ilm_check_missing(d, y ~ x + z)
im <- ilm_impute(d, m = 20)                  # chained equations by default
ilm_mi_pool(im, y ~ x + z)                   # fit each; Rubin's rules
```

`method = "glrm"` imputes categorical columns as categories; `method =
"lowrank"` handles data wider than it is long.

See the *Missing data* vignette.

---

## 6. A causal model, if you have one

Optional, and the stage that decides whether the next one can be read causally.
`ilm_dag()` records what you believe causes what; `ilm_adjust_sets()` finds the
minimal sets of variables that close the back-door paths; `ilm_dag_test()`
checks the conditional independencies the graph implies against the data.

```{r}
g <- ilm_dag("treat -> outcome; conf -> treat; conf -> outcome")
ilm_adjust_sets(g, exposure = "treat", outcome = "outcome")
ilm_dag_test(g, data = d)                    # does the data agree with it?
fit <- ilm_dag_model(g, exposure = "treat", outcome = "outcome", data = d)
```

`ilm_dag_model()` fits the model the adjustment set implies, so the mean
structure is not chosen by looking at p-values.

See the *Causal models* vignette.

---

## 7. The model

One function. The formula is lme4's, the families run from gaussian through
counts, proportions, ordered and unordered categories, and survival times, and
the same object comes back whatever you fit.

```{r}
ilm_model(y ~ x + (1 | subj), data = d, family = "gaussian")
ilm_model(y ~ x + (1 | subj), data = d, family = "multinomial")
ilm_model(y ~ x, data = d, family = "ordinal")
ilm_model(y ~ x, data = d, family = "poisson", ziformula = ~ z)
ilm_model(y ~ x, data = d, family = "beta")
ilm_model(y ~ x + t2(time), data = d, family = "gaussian", ar = ilm_ar1(subj, time))
ilm_iv(y ~ x + w | z + w, data = d)          # an instrument instead
```

Designs the model cannot absorb get their own function: `ilm_did()` for
difference-in-differences, `ilm_rdd()` for a regression discontinuity,
`ilm_design()` for a complex sample.

See the *Regression models* vignette.

---

## 8. Diagnostics

Every check says whether an assumption holds **and names a remedy that exists
in this package**. That is the rule the diagnostics layer is built around; a
check with no remedy is a complaint.

```{r}
ilm_appraise(fit)                            # the whole battery
ilm_check_variance(fit)                      # -> dispformula =
ilm_check_zeros(fit)                         # -> ziformula =
ilm_check_proportional(fit)                  # -> family = "multinomial"
ilm_check_ar(fit)                            # -> ar = ilm_ar1(...)
ilm_variogram(fit)                           # spatial or temporal correlation
ilm_rqr_test(fit)                            # residuals, calibrated by simulation
```

If the variance structure is the worry rather than the mean structure,
`ilm_robust()` gives cluster-robust standard errors without needing to say what
the right structure would be.

---

## 9. What the model says

`ilm_anova()` for which terms matter, `ilm_effects()` for each on its family's
own scale, `ilm_emmeans()` and `ilm_contrast()` for group means and the
comparisons between them, `ilm_ame()` for the average effect of a one-unit
change, and `ilm_plot_model()` for partial dependence.

```{r}
ilm_anova(fit)                               # Type II by default
ilm_effects(fit)                             # odds ratios, rate ratios, ...
em <- ilm_emmeans(fit, "group")
ilm_contrast(em)                             # simultaneous by default
ilm_ame(fit, "x")                            # population-averaged
ilm_plot_model(fit)
```

A ratio from a mixed model is **conditional** on the cluster, and `ilm_effects()`
says so rather than letting it be read as a population quantity.

When the significant term is an interaction between a factor and a *continuous*
predictor -- treatment by time, most often -- a marginal mean cannot take it
apart, because averaging at the mean of the covariate collapses the thing being
asked about. `ilm_trends()` estimates the slope within each level instead, and
answers the two questions separately: whether each slope differs from zero, and
whether the slopes differ from each other.

```{r}
tr <- ilm_trends(fit, "group", var = "time")
tr                                           # is each group's slope non-zero?
ilm_contrast(tr)                             # do the groups' slopes differ?
```

For a mixed model the reference is a large-sample one unless you ask otherwise.
`ilm_denom_df()` supplies Satterthwaite or Kenward-Roger denominator degrees of
freedom, which matter when the clusters are few.

If the design is a factorial or repeated-measures one and you want it reported
the way those designs are usually reported -- F, mean squared error, generalized
eta squared, a sphericity correction -- `ilm_aov_ez()` does the whole of stages
7 to 9 from a description of the columns. See `vignette("anova")`.

---

## 10. Scenario projection

Optional, and the stage a decision usually needs. `ilm_scenario()` sets
predictors to values a stakeholder is considering and reports what the model
says would happen, with intervals and with differences between scenarios.

```{r}
ilm_scenario(fit, dose = c(0, 10, 20), contrast = "first")
```

The default averages over **the observed units**: set the intervention for
everyone, keep each unit's own covariates, predict, average. That is a different
question from predicting for a unit at the average covariate, and under any
non-linear link it is a different answer. It also checks whether each scenario
sits inside the data -- by value, and by combination, since every value can be
ordinary while the combination is one nobody occupies.

---

## 11. Reporting

`ilm_interpret()` writes the findings as prose from deterministic templates, so
the same fit always produces the same words. Causal language appears only when
a DAG has licensed it.

```{r}
ilm_interpret(fit)
```

---

## What to read next

| Vignette | Covers |
|---|---|
| *Introduction to illume* | orientation and one worked example |
| *Regression models* | `ilm_model()` in depth, every family, mixed multinomial |
| *Exploring data* (illumex) | descriptives, plots, bootstrap intervals |
| *Profiling* (illumex) | `ilm_reduce()`, `ilm_cluster()`, `ilm_profile()` |
| *Anomaly detection* (illumex) | `ilm_anomaly()` against `ilm_outliers()`; `ilm_plot_anomaly()` to see whether the flagged rows stand apart |
| *Missing data* | diagnosing it, imputing it, pooling |
| *Causal models* | DAGs, difference-in-differences, discontinuities |
| *Effect size and power* | `ilm_effects()`, `ilm_power()`, `ilm_power_design()`, `ilm_scaffold()`, `ilm_scenario()` |
| *Moderation* | `ilm_moderation()`, `ilm_plot_moderation()` |
